# Pimping your plot



## 

The default formatting used by **ggplot2** is generally fine for exploratory purposes. However, we will often want to change the appearance of a plot once we're ready to include it in a report. The **ggplot2** defaults are carefully chosen to ensure that the information in a plot is easy to discern. These choices are a little unconventional though. For example, published figures usually use a white background. Our aim in this subsection is to learn a little bit about the underlying logic of how to customise **ggplot2**. 

There are a huge number of ways we could customise our example, so I will just show you some of the more common options to illustrate the important ideas.

What do we do if we if we need to change the properties of a geom? We are using the point geom. How might we change the colour or size of points on our plot? You can probably guess: All we need to do is set the appropriate arguments in the `geom_point` function. Let's rerun our example, setting the colour, size and transparency of points:
```{r}
plt <- ggplot(storms, aes(x = pressure, y = wind)) + 
       geom_point(colour = "steelblue", size = 3, alpha = 0.2)
plt
```

The point colour is set with the `colour` argument. There are many ways to specify colours in R, but if you need to specify only one, the simplest is to use a name R recognises. There is nothing special about "steelblue", I just like it. There are over 650 colour names built into R -- the function `colours` will print these to the console for you. The point size is specified with the `size` argument. I increased the point size by assigning this a value of 3. Finally, I made the points somewhat transparent by setting the value of the `alpha` argument (don't ask why it's called this) to be less than 1.

Take another look at that last example. Notice that I split the definition of `plt` over two lines. It will still be treated as a single expression when you do this -- R doesn't much care about white space -- as long as you leave the `+` at the end of each line (__not__ at the beginning). Splitting the different parts of a graphical object definition across lines is generally a good idea, because it makes everything a little more readable. We will do this from now on.

There are other arguments -- such as `fill` and `shape` -- that can be used to adjust the way the points are rendered. The best way to learn how these work is to just experiment with them. However, the key message to take away form this little example is this: if you want to set the properties of __all the elements__ of a particular layer, you do so by specifying the appropriate arguments in the function that defines that layer. You __do not__ need to change the arguments passed to the `ggplot` function.

What else might we like to tweak? Look at the x and y axis labels. These are just the names of the data variables used to define the aesthetic mapping. These labels aren't too bad, but they could be more informative. We know "wind" stands for "wind speed", but someone reading this figure may not realise this immediately. There are also no units -- generally a big no-no for scientific figures. This is how we can manually set the axis labels: 
```{r}
plt <- ggplot(storms, aes(x = pressure, y = wind)) + 
       geom_point(colour = "steelblue", size = 3, alpha = 0.2) + 
       xlab("Atmospheric Pressure (mbar)") + ylab("Wind Speed (mph)")
plt
```

The axes labels are a feature of the whole plot. They do not belong to a particular layer. This is why we don't alter axis labels by passing arguments to the function that built a layer (`geom_point` in this case). Instead, we use the `xlab` and `ylab` functions to set the x and y labels, respectively, using `+` to add them to our graphical object. If you really need to add a title to you graph -- though honestly, when was the last time you read a paper where the figures had titles? -- you can use the `ggtitle` function in the same way.

Let's look at one last way to tweak our plot before finishing this subsection. We are going to adjust the y axis scale. Specifically, we want to increase the number of "guides" -- these are the horizontal lines inside the plot and their accompanying labels. Here is how we place guides at 20, 40, 60, 80, etc, on the y axis:
```{r}
plt <- ggplot(storms, aes(x = pressure, y = wind)) + 
       geom_point(colour = "steelblue", size = 3, alpha = 0.2) + 
       xlab("Atmospheric Pressure (mbar)") + ylab("Wind Speed (mph)") + 
       scale_y_continuous(breaks = seq(20, 160, by = 20)) 
plt
```

What is going on here? We haven't said much about scales up until now. Remember, a scale controls how the data is mapped to the aesthetic attributes, i.e. in this example, how pressure and wind are mapped onto the x and y axes. Here are two important observations: 1) every aesthetic mapping has its own corresponding scale and 2) scales must act on the the whole plot. You should be able to deduce from this that we can't alter scales by manipulating individual layers. Instead, we had to use a function of the general form `scale_XX_YY`. The `XX` bit in the name needs to reference one of the aesthetics, while the `YY` part refers to the kind of scale we want to define. The aesthetic we altered was the y axis, and this is a continuous scale (more about that in the next topic), so we used the `scale_y_continuous` function to tweak the y axis. The `breaks` argument just takes a vector containing a numeric sequence and uses this to specify where the guides should be drawn. Predictably, there is a `scale_x_continuous` function for altering the x axis. 

Scales are one of the trickier aspects of **ggplot2**. There are a lot of them -- type `scale_` at the Console and hit the tab key if you want to see how many -- and each of them can take a variety of arguments. Luckily, the defaults used by **ggplot2** are good enough that we can often arrive at a good plot without having to manipulate.

### Increasing the information density

We introduced this example by saying that we were interested in the relationship between the observation year, the type of storm, wind speed atmospheric pressure, and time of year. So far we have only looked at the middle two. How can we learn about relationship of these two variables to time of year (`seasday`)? We need to include information in the `seasday` variable in our graph. That is, we want to map the `seasday` variable to one of the aesthetics. This means we need to change the way we are using `aes`. One option is to map the `seasday` to the point colours so that the colour of the points describes the time of year. Here is how we specify this:
```{r}
plt <- ggplot(storms, aes(x = pressure, y = wind, colour = seasday)) + 
       geom_point(colour = "steelblue", size = 3, alpha = 0.2) + 
       xlab("Atmospheric Pressure (mbar)") + ylab("Wind Speed (mph)") + 
       scale_y_continuous(breaks = seq(20, 160, by = 20)) 
plt
```

That does not seem to have worked as the result looks the same as our last plot. What went wrong? We are still setting the colour argument inside `geom_point`. When we add a layer, any layer-specific properties that we set will override the aesthetic mappings. We need to remove the `colour = "steelblue"` from inside `aes` to remedy this: 
```{r, fig.width=6}
plt <- ggplot(storms, aes(x = pressure, y = wind, colour = seasday)) + 
       geom_point(size = 3, alpha = 0.2) + 
       xlab("Atmospheric Pressure (mbar)") + ylab("Wind Speed (mph)") + 
       scale_y_continuous(breaks = seq(20, 160, by = 20)) 
plt
```

This plot is a little better. Points are now coloured according to whether they are associated with early (dark blue) or late (light blue) season observations. There is no obvious affect of season, though it is quite hard to be sure because so much overplotting is occurring -- many points occur in the same place. Notice that **ggplot2** automatically adds a legend to the figure to help us interpret it. A colour scale is not much use without a legend.

What else might we do to make the plot a little easier to read? Since wind speed is only measured to the nearest 5 mph, one option is to randomly shuffle the vertical position of each point a little to avoid overplotting. This is called jittering. We can do this by specifying a position adjustment in our layer (remember, position adjustments are part of layers, not the whole graph). Here is how to do this:
```{r, fig.width=6, echo=FALSE}
set.seed(27081975)
```
```{r, fig.width=6}
plt <- ggplot(storms, aes(x = pressure, y = wind, colour = seasday)) + 
       geom_point(alpha = 0.3, size = 3, 
                  position = position_jitter(w = 0, h = 1.3)) + 
       xlab("Atmospheric Pressure (mbar)") + ylab("Wind Speed (mph)") + 
       scale_y_continuous(breaks = seq(20, 160, by = 20)) 
plt
```

We used the `position_jitter` function to associate the necessary information with the position argument. The `w` and `h` arguments of this function specify how much to jitter points in the x and y directions. The resulting plot is a little easier to read (not much to be honest), but we still can't see any obvious effect of season on the relationship between wind speed and pressure. If there is an effect, it is too subtle to detect with just a scatter plot.

There is nothing to  map other variables to different aesthetics if we wanted to try to squeeze even more information into this plot. For example, we could map the storm type variable (`type`) to the point shape if we wanted, using `shape = type` inside `aes`. However, this graph is already a bit too crowded, so let's finish this topic by looking at a different way to examine relationships among variables. 

Remember faceting? This allows us to break up our data set up into subsets according to the unique values of one or two variables, and then produce a separate plot for each subset. What if we want to assess whether or not the wind speed and pressure relationship varies among years? One way to do this is to make a separate scatter plot for each year. Faceting operates on the whole graph, so we can't do it by just changing the properties of a layer. Instead, we have to use a new function to add the faceting information. Here is how we do this using the `facet_wrap` function:
```{r, fig.height=4.3, fig.width=6.7}
plt <- ggplot(storms, aes(x = pressure, y = wind, colour = seasday)) + 
       geom_point(alpha = 0.3, size = 2, 
                  position = position_jitter(w = 0, h = 1.3)) + 
       xlab("Atmospheric Pressure (mbar)") + ylab("Wind Speed (mph)") + 
       scale_y_continuous(breaks = seq(20, 160, by = 20)) + 
       facet_wrap(~ year, nrow = 2, ncol = 3)
plt
```

The first argument (`~ year`) says to break up the data according to the values of `year`. You need to the `~` at the front by the way. We'll learn what that is doing next year, but for now, take my word for it -- you need to include it for the faceting to work. The `nrow` and `ncol` arguments just specify how to split the panels across rows and columns of the resulting plot. Notice that each panel shares the same scales. This makes it easy to compare information across them. The plot indicates that the wind speed / pressure relationship is more or less invariant across years, and that perhaps 1997 and 2000 were not such bad storm years compared to the others. This probably shouldn't surprise us. Although the occurrence of storms is somewhat stochastic, the laws of atmospheric physics do not change from one year to the next.





