---
title: "Packages and help"
author: "Bethan Hindle"
date: "15 February 2018"
output:
  ioslides_presentation:
    css: extras.css
    smaller: no
    wide: yes
highlighter: pygments
---

```{r, echo=FALSE}
showAnswers <- TRUE
```

## Before we start...

DO THIS NOW!

- Go to the APS 135 MOLE site and open today's lecture ("Presentation 2"). You can navigate through this with the arrow keys.

- Open up RStudio from the Programs Menu on your PC. Grab a TA for help if there seems to be a problem.

- Open a new script.

- Get your ucards out to be scanned.

Refresher exercises
====================================================================

## Exercise {.build}

<div id="exercise-box">
__Vectorised operations__
<div id="box-text">
Use the `c` function to make a pair of numeric vectors containing the two following sets of numbers:

a. 7, 2, 5, 3, 1   
b. 1, 4, 8, 3, 9

Assign them the names `A` and `B`. Write some R code to determine the result of `A + B` and `A + B + 6`. 

Look carefully at the results. What happens when we add two numeric vectors together? What happens when we add a single number to a vector? 

Now make a new numeric vector containing the following set of numbers (call it `X`):

1, 2

Add this to the `A` vector. What happens when you add numeric vectors with different lengths?
</div>
</div>


## Exercise {.build}

<div>
```{r, include=showAnswers}
# make `A`
A <- c(7, 2, 5, 3, 1)
A
# make `B`
B <- c(1, 4, 8, 3, 9)
B
```
</div>

## Exercise {.build}

<div>
```{r, include = showAnswers}
A + B
A + B + 6
```
</div>

```{r, include = showAnswers}
X <- 1:2
A + X
```

## Exercise {.build}

<div id="exercise-box">
__Extracting elements from a vector__
<div id="box-text">

Create a vector containing the letters of the alphabet in reverse order. 

Extension: Imagine that you want to access the last element of a vector, but you don't know how long it is (i.e. you can't use `LETTERS[26]`). How could you do it? 

</div>
</div>


```{r, include=showAnswers}
LETTERS[26:1]
```

## Exercise

```{r, include=showAnswers}
length(LETTERS)
LETTERS[length(LETTERS)]
```

One last bit of the basics... | Logical comparisons
====================================================================

## Comparing vectors {.build .smaller}

<div>
We use __relational operators__ to compare the elements of numeric vectors:

- `x < y `: is x less than y?
- `x > y `: is x greater than y?
- `x <= y`: is x less than or equal to y?
- `x >= y`: is x greater than or equal to y?
- `x == y`: is x equal to y?
- `x != y`: is x not equal to y?

These operators produce something called a __logical vector__ (a vector of TRUE and FALSE values---nothing else allowed)
</div>

<div>
__Example__
```{r}
x <- 1:9
y <- rep(5, 9)
x == y
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Using relational operators__
<div id="box-text">
Make two vectors named `num1` and `num2`, where `num1` is a sequence from 0 to 20, and `num2` is a sequence from 25 to 5. Once you have made these print them to screen, then experiment with `<`, `>`, `<=`, `>=`, `==` and `!=` to convince yourself you understand how they work.
</div>
</div>

```{r, echo=FALSE}
num1 <- 0:20
num2 <- 5:25
```

<div>
```{r, include=showAnswers}
num1
num2

```
</div>

## Exercise {.build}

<div>
```{r, include=showAnswers}
num1
num2
```
</div>

```{r, include=showAnswers}
num1 == num2
num1 != num2
```

## Exercise {.build}

<div>
```{r, include=showAnswers}
num1
num2
```
</div>

```{r, include=showAnswers}
num1 <  num2
num1 <= num2
```

## Combining comparisons {.build .smaller}

<div>
We use __logical operators__ to combine logical statements:

- `!x `: __negation__ = `TRUE` becomes `FALSE`, `FALSE` becomes `TRUE`
- `x & y `: __AND__  = `TRUE` if __both__ x and y are `TRUE`, `FALSE` otherwise
- `x | y`: __OR__  = `TRUE` if __either__ x and y are `TRUE`, `FALSE` otherwise

These operators produce another __logical vector__ (a vector of TRUE / FALSE values)
</div>

<div>
__Example__
```{r}
x <- c(TRUE, TRUE, FALSE, FALSE)
y <- c(FALSE, TRUE, FALSE, TRUE)
x & y
x | y
```
</div>

R packages | A great source of confusion 
====================================================================

## A (very) quick overview {.build}

<div>
Packages extend what base R can do. Two things have to happen before you can use a package:

1. You have to **install** it. This is a "do once" operation. After a package is installed you can close down and reopen R and it is still installed.

2. You have to **load and attach** it. This is a "do every time" operation. You have to load and attach the package every time you start a new session (= open R).
</div>

<hr>

<div>
A mobile phone App analogy: 

- installing a package <=> getting an App from an App store

- loading and attaching a package <=> opening the App on your phone
</div>

## How do we access packages?

1. We **install** a package with the `install.packages` function

2. We **load and attach** an installed package with the `library` function

**TIPS**

- Do not use `install.packages` inside your scripts! You don't want to reinstall a package every time you plan to use it. 

- You can also install a package via the Packages tab in RStudio. It is fine to use this if you don't like using `install.packages`.

## Exercise {.smaller}

<div id="exercise-box">
__Working with packages__
<div id="box-text">
We want to use the `fortune` function (singular!) that resides in the **fortunes** package (plural!). Do the following, in the order given, working directly in the Console:

1. Try to use the `fortune` function, i.e. type `fortune()` straight into the Console. Does this work or do you see an error?

2. Try to use `library` to load and attach the **fortunes** package. Does this work or do you see an error?

3. Look in the **Packages** tab in RStudio. Can you see anything called fortunes? Now install the **fortunes** package with `install.packages`. Look for it the **Packages** tab again. 

4. Try again to use the `fortune` function. Does this work? If not, why not?

5. Use `library` to load and attach the **fortunes** package (plural!). What happens?

6. Try again to use the `fortune` function. Does this work or do you see an error?
</div>
</div>


## Exercise {.smaller .build}

<div id="exercise-box">
__Working with packages (still)__
<div id="box-text">

Now try the following:

1. Close down Rstudio completely then open it again.

2. Try to use the `fortune` function, i.e. type `fortune()` straight into the Console. Does this work or do you see an error?

3. Use `library` to load and attach the **fortunes** package (plural!). What happens?

4. Try again to use the `fortune` function. Does this work or do you see an error?

</div>
</div>

## Exercise {.small}

In the next few weeks we're going to be using two packages: `dplyr` and `ggplot2`.
The TAs installed these packages on a lot of your computers last week.   
<br>
Check whether they're installed now using the following code:

```{r, warning = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
```

If you get an error message (NOT a warning) - stick your hand up.

Getting help | Learning about a function
======================================================================================

## Help files

Do the following:

```{r, eval=FALSE}
?round
```

Help files are really **documentation** files:

- Description
- Usage
- Arguments
- Details
- References
- Examples

Let's step though the help for `round`...

## Exercise {.smaller .build}

<div id="exercise-box">
__Using the help files__
<div id="box-text">

Use the `rep` function to create the following three sequences:
```{r, echo=FALSE}
rep(1:6, each = 3)
rep(c(2, 5), times = 4)
rep(3:6, each = 2, times = 3)
```

</div>
</div>

## Exercise {.build}

```{r, include=showAnswers}
rep(1:6, each = 3)
rep(c(2, 5), times = 4)
rep(3:6, each = 2, times = 3)
```

## Exercise {.smaller .build}

<div id="exercise-box">
__Using the help files__
<div id="box-text">

Use the `seq` function to create the following numeric vector. 
```{r, echo=FALSE}
seq(0, 1, by = 0.1)
```
Assign the vector to a name of your choice.   
Change the 6th element to have a value of 14. 
</div>
</div>

```{r, include=showAnswers}
myseq <- seq(0, 1, by = 0.1)
myseq
```

```{r, include=showAnswers}
myseq[6] <- 14
myseq
```

## What if you don't know the function name?
Do the following:

```{r, eval=FALSE}
??round
```

## Other sources of help {.smaller .build}

The internet is packed full of helpful resources:

- **Google** -- this might be obvious, but you have to work on your "Google-fu" to find what you want. The key is to pick the right search terms:

    - include the function or package name
    
    - include the letter R!
    
    - be selective about which sites you look at...

- **Stack Overflow** -- probably the best question and answer site for computing stuff. People can be a bit blunt though:

    - check whether your question has been asked before (Google)
    
    - read the community guidelines
    
    - provide example code with your questions

## Other sources of help {.build}

- **Each other**
  
- **The TAs**
  
- **Me** - email (bjhindle1@sheffield.ac.uk) to arrange a meeting - if none of the above have helped


## Changing the appearance of RStudio

Navigate through RStudio's menus to Tools > Global Options

Click on Appearance

Experiment with the different themes

## Useful keyboard shortcuts

Move cursor to script:    Ctrl + 1  
Move cursor to console:   Ctrl + 2
Navigate command history: Up/down  
Assignment operator:      Alt + -

## And finally...

- There will **not** be a lecture tomorrow morning.

- Keep working through the online coursebook

- Provide feedback on the course so far at: https://goo.gl/forms/x2x5JZdhygDti09q2

- Next week we'll actually start using some data
