---
title: Data frames and data wranging with dplyr
author: Dylan Childs
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation:
    css: extras.css
    smaller: no
    wide: yes
highlighter: pygments
---

```{r, echo=FALSE, warning = FALSE, message = FALSE}
showAnswers <- TRUE
library(dplyr)
ca_storms <- nasaweather::storms
```

## Before we start...

DO THIS NOW!

- Go to the course MOLE site and open today's lecture ("Block 2"). 

- Open up RStudio from the Programs Menu on your PC.

- Start a new script.

- Get your ucards out to be scanned.

Feedback
======================================================================================

Data frames | A place for your data
======================================================================================

## Two views of data frames {.build}

The first view: __Data frames are a collection of vectors__. These may be different types (e.g. numeric, character, logical), but they must all be the same length.

<div>
Here is how to make one "by hand" using the `data.frame` function:

```{r}
mydf <- data.frame(Var1 = 1:4, Var2 = LETTERS[1:4], Var3 = c(1,2,4,8))
mydf
```

Data frames are central to data analysis in R. You can think of each vector as corresponding to a __statistical variable__ that we want to analyse (more on that in the course book).
</div>

## Selecting vectors from data frames {.build}

We can extract a vector from a data frame in a few different ways:

<div>
- Using the name and the `$` operator
```{r}
mydf $ Var1
```
- Using the name and the `[[ ]]` operator
- Using the position and the `[[ ]]` operator
</div>

## Selecting vectors from data frames

We can extract a vector from a data frame in a few different ways:

- Using the name and the `$` operator
- Using the name and the `[[ ]]` operator
```{r}
mydf[["Var1"]] # <- WE HAVE TO USE QUOTES!
```
- Using the position and the `[[ ]]` operator


## Selecting vectors from data frames

We can extract a vector from a data frame in a few different ways:

- Using the name and the `$` operator
- Using the name and the `[[ ]]` operator
- Using the position and the `[[ ]]` operator
```{r}
mydf[[1]]
```

<br><br><br><hr>

N.B. We will only use the first version in this course. Remember what `$` does. You **will** need to use it.

## Exercise {.build}

<div id="exercise-box">
__Extracting vectors from data frames__
<div id="box-text">
R has many datasets built into it, stored as data frames. One of these is the famous `iris` data. Type this into your Console: `View(iris)`. How many vectors (i.e. columns) are stored in `iris`? Extract the one named `Sepal.Width`.
</div>
</div>

<div>
```{r, include=showAnswers}
iris$Sepal.Width
```
</div>

## Two views of data frames {.build}

The second view: __Data frames are table-like objects__. Take another look at our simple data frame:

```{r}
mydf
```

This clearly has rows and columns. We can extract elements of the data frame using the square brackets `[ ]`:

```{r}
mydf[2, 3]
```

__[Row #, Col #]__: This extracted the element in the second row and third column. 

## Exercise {.build}

<div id="exercise-box">
__Extracting elements from data frames__
<div id="box-text">
See if you can use the `[,]` construct to get the elements in rows 2 to 3 __and__ columns 1 to 4 from the `iris` data. Hint: you will need to use the `:` operator, twice inside `[,]` 
</div>
</div>

<div>
```{r, include=showAnswers}
iris[2:3, 1:4]
```
</div>

__Note__: this produced a new data frame.

## Exercise {.build .smaller}

<div id="exercise-box">
__Examining a data frame__
<div id="box-text">
We've seen that `View(iris)` can be used to examine a data frame in a spreadsheet like view. There are many other functions that allow us to examine the structure of a data frame. Using the `iris` data, experiment with the `head`, `tail`, and `str` functions to see what they do.
</div>
</div>

<div>
```{r, include=showAnswers}
head(iris, n = 2)
```
</div>

<div>
```{r, include=showAnswers}
str(iris)
```
</div>

The working directory | Where R "looks" for files
======================================================================================

## What is the working directory? {.build}

A quick overview of what it is and how to work with it...

- The **working directory** is a default location where R looks for files you want to use. It is just a folder on your computer.

- If you don't set the working directory, R will do it for you. It will seldom choose the exact folder you want to use.

- YOU should set the working directory EVERY time you start R/RStudio. Setting your working directory should be the first thing you do.

- Don't do this using R code in your script. Instead, do it via the Console (hard and error prone) or using the RStudio menu system (easy and reliable).

<div>
**Set your working directory now using...** 
**`Session > Set Working Directory > Choose Directory...`**
</div>


Reading in data | A great source of confusion
====================================================================

## We need some data....

**1. Grab the "iris_morphology.xlsx" file from MOLE**:

(APS135 or APS118 > Exploratory Data Analysis with R  > Data files needed for R course (semester 2))

**2. Save this anywhere you like**

Make sure you remember where you put it!

**3. Open up your copy of "iris_morphology.xlsx" in Excel**

How many columns ("variables") and rows ("observations") are there?


## Getting data from Excel into R {.build .smaller}

Essentially a two step process

<div>
**STEP 1. Export data from Excel to a CSV file**:

- Open the "iris_morphology.xlsx" spreadsheet up in Excel

- In Excel, go to `Save As...` and choose the `Comma Separated Values (csv)` option

- Keep the default file name ("iris_morphology.csv"), but make sure the file is saved **in your working directory** (you set this earlier)
</div>
<div>
__STEP 2. Now import the CSV file into using RStudio:__

- Include the following line in your script, and send it to the Console

```{r, eval = FALSE}
my_iris <- read.csv("iris_morphology.csv", stringsAsFactors = FALSE)
my_iris
```

- What happened? Did you see any error messages? If you did, check your spelling and make sure your working directory is set to the right place.
</div>

## The other way...

__This is how to import data in a CSV file using RStudio:__

- Select `Import Dataset > From Text File...`

- Select the file your saved and click Open

- Enter a name (no spaces allowed) and click Import

**Some advice: Don't use this method!!!** You have to remember do it everytime you want to run your script. This is very error prone.


Overview of **dplyr** and `tbl` | Getting to grips with the basics 
====================================================================

## Why use dplyr?

**dplyr** implements a __grammar of data manipulation__ to enable you manipulate data and summarise the information in a data set (e.g. group means).

__Advantages of using **dplyr**__
 
- Provides a consistent and easy to understand framework for data manipulation
- Designed to work well with the `ggplot2` plotting system
- Fast compared to many base R functions 
- Allows you to work with data stored in many ways (e.g. in a database)

## Five key verbs {.build} 

<div>
__**dplyr** has five main "verbs" (i.e. functions):__

- `select`: Extract a subset of variables
- `filter`: Extract a subset of rows
- `arrange`: Reorder rows 
- `mutate`: Construct new variables 
- `summarize`: Calculate information about groups

(+ a few other important functions, e.g. `slice`, `rename`, `transmute`, and `group_by`)
</div>

<div>
<br>

## Five key verbs
__It is helpful to classify the verbs according to what they work on:__

- observations (rows): `filter` & `slice` & `arrange`
- variables (columns): `select` & `rename` & `mutate`
- summarise: `summarize` (or `summarise`)

(This classification only works if your data are __tidy__. Make sure you read about this)
</div>

## Before we start {.build .smaller}

Required packages and data:

- **dplyr**: obviously
- `CA_STORMS.csv`: climate data for Central America (on MOLE)

<div>
**1. Load the __dplyr__ package**

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```

This should already be installed. If you see errors, you may have to install it, e.g.

```{r, eval=FALSE}
install.packages("dplyr")
```
</div>

<div>
__2. Read in the storms data__

Download the `CA_STORMS.csv` data from MOLE. Set your working directory, and then use `read.csv` to read the data in (remember to use `stringsAsFactors = FALSE`). Assign the resulting data frame the name `ca_storms`.
</div>


__3. Use `View` to examine the `ca_storms` data__

## Making and using `tbl` objects  {.smaller}

A `tbl` object (pronounced "tibble") is essentially a special kind of data frame. It works the same as a data frame, but with a few small differences... e.g. compact printing:

```{r}
tbl_df(ca_storms)
```

## Convert `ca_storms` and `iris` to tibbles now

Next in your script, add these lines to convert `ca_storms` and `iris` to tibbles...

```{r}
ca_storms <- tbl_df(ca_storms)
iris   <- tbl_df(iris)
```

You don't have to do this (**dplyr** is fine with normal data frames), but it will ensure your output matches the presentation.

## Looking at your `tbl` (or `data.frame`) 

In addition to printing a `tbl` or `data.frame`, we can use the `glimpse` function with these objects to obtain different summary information about variables:

```{r}
glimpse(ca_storms)
```

Glimpse tells us what variables are in `storm` as well as the type of each variable.

Using `select` | Extracting a subset of variables 
====================================================================

## Basic usage

We use `select` to extract a subset of variables for further analysis. Using `select` looks like this:

```{r, eval=FALSE}
select(data, Variable1, Variable2, ...)
```

__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `VariableX`: names of variables in `data`

## Exercise {.build}

<div id="exercise-box">
__Selecting two variables__
<div id="box-text">
Use the `select` function with the `ca_storms` data set to make a new data set containing only `name` and `year`. Assign this new data set a name, and then check that it contains the right variables using the `glimpse` function.
</div>
</div>

```{r, include=showAnswers}
ca_storms_simple <- select(ca_storms, name, year)
glimpse(ca_storms_simple)
```

## Selecting & dropping variable ranges {.build}

The `select` function makes selecting/removing groups of variables easy:

- Use `:` to select a sequence of variables 
- Use `-` to drop a sequence of variables 

The sequence can be specified using numbers (for position) or names.

<div>
__Usage:__

```{r, eval=FALSE}
# a range of variable to keep
select(data, Variable1:Variable5)
# a range of variable to drop
select(data, -(Variable1:Variable5))
```
</div>

<div>
## Example:

```{r}
iris_fewer <- select(iris, Petal.Length:Species)
iris_fewer
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Selecting a range of variables__
<div id="box-text">
Use the `select` function with the `ca_storms` data set to select __just__ the variables `name`, `year` and `month` variables.
</div>
</div>

```{r, include=showAnswers}
ca_storms_fewer <- select(ca_storms, name:month)
glimpse(ca_storms_fewer)
```

```{r, include=showAnswers}
# alternatively
x <- select(ca_storms, -(day:seasday))
```

## Helper functions {.build .smaller}

There are several helper functions that work with `select` to simplify common variable selection tasks:

- `starts_with("xyz")`: every name that starts with `"xyz"`
- `ends_with("xyz")`: every name that ends with `"xyz"`
- `contains("xyz")`: every name that contains `"xyz"`
- `matches("xyz")`: every name that matches `"xyz"`
- `one_of(names)`: every name that appears in `names` (character vector).

<div>
__Usage:__

```{r, eval=FALSE}
select(data, help_func("xyz"))
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_petal <- select(iris, starts_with("Petal"))
glimpse(iris_petal)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Using select with helper functions__
<div id="box-text">
Use the `select` function with the `ca_storms` data set to create a new data set containing just the `lat` and `long` variables. Do this using the `starts_with` helper function inside select. 
</div>
</div>

```{r, include=showAnswers}
ca_storms_fewer <- select(ca_storms, starts_with("l"))
glimpse(ca_storms_fewer)
```

Using `select` and `rename` | Renaming variables
====================================================================

## Renaming while selecting {.build}

We can use `select` to rename variables as we select them using the `newName = varName` construct.

<div>
__Usage:__

```{r, eval=FALSE}
select(data, newName1 = Var1, newName2 = Var2, ...)
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_select <- select(iris, PetalLength = Petal.Length)
glimpse(iris_select)
```
</div>

## Renaming without selecting {.build}

Use `rename` to rename variables while keeping all variables using the `newName = varName` construct.

<div>
__Usage:__

```{r, eval=FALSE}
rename(data, newName1 = Var1, newName2 = Var2, ...)
```
</div>

<div>
__Example:__

```{r}
iris_renamed <- rename(iris, 
                       PetalLength = Petal.Length, 
                       PetalWidth  = Petal.Width)
glimpse(iris_renamed)
```
</div>


