---
title: More data wrangling with dplyr
author: Bethan Hindle
date: March 1, 2018
highlighter: pygments
output: 
  ioslides_presentation:
    css: extras.css
    widescreen: true
    smaller: true
---

```{r, echo=FALSE, warning = FALSE, message = FALSE}
showAnswers <- TRUE
library(dplyr)
library(nasaweather)
```

## Before we start {.build .smaller}

Required packages and data:

- `dplyr`
- `storms.csv`: climate data for Central America (on MOLE)

**1. Load the `dplyr` package**

```{r, message=FALSE}
library("dplyr")
```

__2. Read in the storms data__

Download the `storms.csv` data from MOLE (if you didn't already do this last week). Set your working directory to the correct location and then use `read.csv` to read the data in (remember to use `stringsAsFactors = FALSE`), giving it the name `storms`.

__3. Convert the storms dataframe to a tbl__

Remember that we can use the `tbl_df` function to convert a data frame to a tbl. This is basically a special kind of data frame that has a few advantages such as compact printing.

Using `select` and `rename` | Renaming variables
====================================================================

## Renaming while selecting {.build}

We can use `select` to rename variables as we select them using the `newName = varName` construct.

<div>
__Usage:__

```{r, eval=FALSE}
select(data, newName1 = Var1, newName2 = Var2, ...)
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_select <- select(iris, PetalLength = Petal.Length)
glimpse(iris_select)
```
</div>

## Renaming without selecting {.build}

Use `rename` to rename variables while keeping all variables using the `newName = varName` construct.

<div>
__Usage:__

```{r, eval=FALSE}
rename(data, newName1 = Var1, newName2 = Var2, ...)
```
</div>

<div>
__Example:__

```{r}
iris_renamed <- rename(iris, 
                       PetalLength = Petal.Length, 
                       PetalWidth  = Petal.Width)
glimpse(iris_renamed)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Renaming__
<div id="box-text">
Extract just the `lat` and `long` columns from the storms data set and rename these as `latitude` and `longitude` respectively.

</div>
</div>

```{r, include=showAnswers}
storms_renamed <- select(storms, latitude = lat, longitude = long)
glimpse(storms_renamed)
```

Using `mutate` | Making new variables 
====================================================================

## Basic usage {.build}

<div>
We use `mutate` to __add variables__ for further analysis. This is how we use `mutate`:

```{r, eval=FALSE}
mutate(data, NewVar = <expression>, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `NewVar`: name of a new variable to create
- `<expression>`: an R expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` which appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`.

- You may use more than one `NewVar = <expression>` at a time if you need to construct several new variables.
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
Use the `mutate` function with the `iris` data set to make a new variable which is the petal area, $(Area = Length \times Width)$.
</div>
</div>

```{r, include=showAnswers}
iris_area <- mutate(iris, Petal.Area = Petal.Length * Petal.Width)
glimpse(iris_area)
```

## Multiple calculations {.build}

<div>
We can add more than one variable at a time using `mutate`. Each new variable can also use one or more variables created in a previous step.
</div>

<div>
__Usage:__

```{r, eval=FALSE}
mutate(data, NewVar1 = <expression1>, 
             NewVar2 = <expression2 using NewVar1>)
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_new_vars <- 
  mutate(iris, Sepal.Eccentricity = Sepal.Length / Sepal.Width,
               Petal.Eccentricity = Petal.Length / Petal.Width,
               Eccentricity.Diff  = Sepal.Eccentricity - Petal.Eccentricity)
glimpse(iris_new_vars)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Making several new variables__
<div id="box-text">
Use the `mutate` function with the `iris` data set to make two new area variables, one for petal and one for sepal. Create a third variable which is the ratio of the petal and sepal areas. 
</div>
Do all of this in one call to `mutate`, i.e. use `mutate` only once to do all of this.
</div>
</div>

```{r, include=showAnswers}
iris_ratio <- mutate(iris, Sepal.Area = Sepal.Length * Sepal.Width,
                           Petal.Area = Petal.Length * Petal.Width,
                           PS.Area.Ratio = Petal.Area / Sepal.Area)
glimpse(iris_ratio)
```

Using `filter` | Selecting subsets of observations 
====================================================================

## Basic usage {.build}

<div>
We use `filter` to select a subset of rows for further analysis, based on the result(s) of one or more logical comparisons. Using `filter` looks like this:

```{r, eval=FALSE}
filter(data, <expression>)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `<expression>`: an R expression that implements a logical comparison using variables in `data`
</div>

<div>
__Comments__

- The `<expression>` can be any valid R expression that uses variables in `data` and returns a logical vector of `TRUE` / `FALSE` values.
- The `<expression>` typically uses a combination of __relational__ (e.g. `<` and `==`) and __logical__ (e.g. `&` and `|`) operators
</div>

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on one variable__
<div id="box-text">
Use the `filter` function with the `storms` data set to create a new data set containing just the observations associated with storms classified as Hurricanes.

Hint: use `glimpse` to remind yourself of the variable names in `storms`. You need to work out which one contains information about the storm category.
</div>
</div>

```{r, include=showAnswers}
filter(storms, type == "Hurricane")
```

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on more than one variable__
<div id="box-text">
Repeat the last exercise, but now extract the observations associated with Hurricanes that took place in 1997 or later.
</div>
</div>

```{r, include=showAnswers}
filter(storms, 
       type == "Hurricane" & year >= 1997) 

# or use: filter(storms, type == "Hurricane", year >= 1997)
```


Using `arrange` | Reordering observations 
====================================================================

## Basic usage {.build}

<div>
We use `arrange` to reorder the rows of our data set. This can help us see associations among our variables. Using `arrange` looks like this:

```{r, eval=FALSE}
arrange(data, Variable1, Variable2, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `VariableX`: names of variables in `data`
</div>

<div>
__Comments__

- The order of sorting corresponds to the order that variables appear in `arrange`, meaning `data` is sorted according to `Variable1`, then `Variable2`, then `Variable3`, etc
- The sort order is from smallest to largest (ascending). If you want to reverse the sort order to go from largest to smallest (descending) use `desc(VariableX)`.
</div>

## Exercise {.build}

<div id="exercise-box">
__Reording observations__
<div id="box-text">
Use the `arrange` function to reorder the observations in the `storms` data set, according to the pressure variable. Store the resulting data set and use then use the `View` function to examine it. What can you say about the association between atmospheric pressure and storm category?
</div>
</div>

```{r, include=showAnswers}
storm.sort <- arrange(storms, pressure)
```

```{r, eval=FALSE}
View(storm.sort)
```

Using `summarise` | Calculating summaries of variables 
====================================================================

## Basic usage {.build .smaller}

<div>
We use summarise to __calculate summary statistics__ for further analysis. This is how to use `summarise`:

```{r, eval=FALSE}
summarise(data, SummaryVar = <expression>, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `SummaryVar`: name of your summary variable
- `<expression>`: an R expression that references variables in `data` and returns to __a single value__
</div>

<div>
__Comments__

- The `<expression>` which appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`. However, `<expression>` should return __a single value__.
- Although `summarise` looks a little like `mutate`, it is designed to construct a completely new dataset containing summaries of one or more variables.
- You may use more than one `SummaryStat = <expression>` at a time if you need to construct several summaries.
</div>

## Exercise {.build}

<div id="exercise-box">
__Calculating the mean of two variables__
<div id="box-text">
Use the `summarise` function with the `iris` dataset to calculate the mean sepal length and the mean sepal width.

Hint: You need to work out which R function calculates a mean. The clue is in the name.
</div>
</div>

```{r, include=showAnswers}
summarise(iris, 
          mean_sl = mean(Sepal.Length), 
          mean_sw = mean(Sepal.Width))
```

## Exercise {.build}

<div id="exercise-box">
__Calculating a more complex summary of a variable__
<div id="box-text">
Use the `summarise` function with the `iris` dataset to calculate the mean __area__ of sepals.
</div>
</div>

```{r, include=showAnswers}
summarise(iris, mean_sl = mean(Sepal.Length * Sepal.Width))
summarise(iris, mean_sl = mean(Sepal.Length) * mean(Sepal.Width))
```

Which one is right?

Using `group_by` | Making summaries for groups of observations
====================================================================

## Basic usage{.build .smaller}

<div>
We use `group_by` to add grouping information to a data frame or tibble. This is how we use `group_by`:

```{r, eval=FALSE}
group_by(data, GroupVar1, GroupVar2, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `GroupVar`: name of grouping variable(s) 
</div>

<div>
__Comments__

- The `group_by` does not do anything other than add grouping information to a `tbl`. It is only useful when used with `summarise` or `mutate`.
- Using `group_by` with `summarise` enables us to calculate numerical summaries on a _per group_ basis.
- Using `group_by` with `mutate` enables us to add variables on a _per group_ basis.
</div>

## Exercise {.build .smaller}

<div id="exercise-box">
__Using `group_by` to calculate group-specific means__
<div id="box-text">
Use the `group_by` function and the `summarise` functions with the `storms` dataset to calculate the mean wind speed associated with each storm __type__.

Hint: This is a two step exercise: 1) Use `group_by` to add some information to storms, remembering to assign the result a name; 2) These use `summarise` on this new dataset.
</div>
</div>

```{r, include=showAnswers}
# 1. make a grouped tibble
storms_grouped <- group_by(storms, type)
# 2. use summarise on the grouped data
summarise(storms_grouped, mean_wind = mean(wind))
```

## Exercise {.build .smaller}

<div id="exercise-box">
__Using `group_by` to group by more than one variable__
<div id="box-text">
Use the `group_by` and `summarise` functions with the `storms` dataset to calculate the mean and maximum wind speed associated with **each combination** of month and year. Assign the result a name and then use `View` to examine it. Which month in which year saw the largest maximum wind speed? 

Hint: You can guess the names of the two functions that calculate the mean and max from a numeric vector.
</div>
</div>

```{r, include=showAnswers}
# 1. make a grouped tibble
storms_grouped <- group_by(storms, year, month)
# 2. use summarise on the grouped data
summarise(storms_grouped, mean_speed = mean(wind), max_speed = max(wind))
```

## Exercise {.build .smaller}

<div id="exercise-box">
__Using `group_by` and `mutate`__
<div id="box-text">
The `group_by` function works with any **dplyr** verb that operates on variables (columns). Use the `group_by` and `mutate` functions with the `iris` dataset to calculate a "mean centred" version of sepal length. A centred variable is just one that has had its overall mean subtracted from every value. 

Do you understand the different behaviour of `summarise` and `mutate` when used alongside `group_by`?
</div>
</div>

```{r, include=showAnswers}
# 1. group iris by species identity
iris_grouped <- group_by(iris, Species)
# 2. use mutate on the grouped data
mutate(iris_grouped, sl_centred = Sepal.Length - mean(Sepal.Length))
```

## What happens next?

<br><br>

1. __Lecture__ tomorrow.

2. Keep reading the online course book!

3. Email me if you've stuck/struggling (bjhindle1@sheffield.ac.uk)

4. We'll start learning to make plots next week

5. Fill in a feedback form https://goo.gl/forms/NcoUP5KHlwacUAsj2













