---
title: Final dplyr and ggplot introduction
author: Bethan Hindle
date: March 8th, 2018
highlighter: pygments
output: 
  ioslides_presentation:
    css: extras.css
    widescreen: true
    smaller: false
---

```{r, include = FALSE}
showAnswers <- TRUE
library(dplyr)
library(ggplot2)
library(nasaweather)
library(grid)
library(knitr)

opts_chunk$set(fig.width=4.5, fig.height=3.3, warning=FALSE, message=FALSE, 
               dev.args = list(bg = 'transparent'), fig.align='center')

my_theme <- theme_gray() + theme(
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(vjust=-0.5),
  axis.title.y = element_text(vjust=+1.5)
  )
theme_set(my_theme)
```

## Before we start {.build .smaller}

Required packages and data:

- `dplyr`: obviously
- `ggplot2`: new package
- `storms.csv`: climate data for Central America (on MOLE)

**1. Load the `ggplot2` package**

```{r, message=FALSE}
library("ggplot2")
```

This should already be installed. If you see errors, you may have to install it, e.g.

```{r, eval=FALSE}
install.packages("ggplot2")
```

Do the same for `dplyr` (you should NOT need to install it first)

__2. Read in the storms data__

Set your working directory to the correct location and then use `read.csv` to read the `storms.csv` data in (remember to use `stringsAsFactors = FALSE`), giving it the name `storms`.


Using `%>%` | piping or chaining
====================================================================

## Motivating example {.build .smaller}

We often need to perform a sequence of calculations on a vector or data frame. We do this by applying a series of function in sequence. Here are two ways to do this:

<div>
**Method 1: Store intermediate results...** 


```{r}
x <- 10
x <- sqrt(x)
x <- exp(x)
round(x, 2)
```
</div>

<div>
**Method 2: Use function nesting...** 

```{r}
round(exp(sqrt(10)), 2)
```
</div>

These do the same thing. Method 1 is easy to read, but is very verbose. Method 2 is concise, but not at all easy to read.

## A third way, using `%>%`...{.build}

The `dplyr` package includes a special operator, called "the pipe". It looks like this: `%>%`. This allows us to avoid storing intermediate results (method 1), while reading a sequence of function calls from left to right. For example:

```{r}
10 %>% sqrt(.) %>% exp(.) %>% round(., 2)
```

<div>
Or equivalently, and even simpler...

```{r}
10 %>% sqrt() %>% exp() %>% round(2)
```
</div>

## Why is this handy? {.build .smaller}

We can use `%>%` with any function we like. Look at this example of a two-step `group_by` and `summarise` operation with `iris`...

<div>
```{r, eval=FALSE}
# method 1
iris.grouped <- group_by(iris, Species)
summarise(iris.grouped, mean.SL = mean(Sepal.Length))

# method 2
summarise(group_by(iris, Species), mean.SL = mean(Sepal.Length))
```
</div>

<div>
The "piped" equivalent is more natural to read...

```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(mean.SL = mean(Sepal.Length))
```
</div>

## Exercise {.build .smaller}

<div id="exercise-box">
__Using `group_by` to calculate group specific summaries__
<div id="box-text">
Last week we used `group_by` and `summarise` with the `storms` data set to calculate the mean wind speed associated with each type of storm. Repeat this exercise now using the pipe. 
</div>
</div>

```{r, include=showAnswers}
storms %>% 
  group_by(type) %>%
  summarise(mean_speed = mean(wind))
```


Overview of `ggplot2` | The grammar of graphics
====================================================================

## Why use `ggplot2`? {.build .smaller}

<div>
Roughly speaking, there are three commonly used plotting frameworks in R. 

- base graphics: available as soon as you open R
- `lattice` package
- `ggplot2` package
</div>

<div>
__Advantages__ of using `ggplot2`
 
- Consistent and intuitive framework for plotting
- Flexible enough to make every plot you will need
- Works well with `dplyr`
</div>

<div>
__Disadvantages__ of using `ggplot2`

- You have to learn "the grammar" to use it well
- Vast package, can be intimidating
- More than one way to do things
</div>

## Key concepts {.build}

You need to wrap your head around a few ideas to start using `ggplot2` effectively:

>- __layers__: We build ggplot2 objects by adding one or more layers together. We can do this in a stepwise way, only plotting the object when we are ready. Each layer needs things like data, aesthetics, geometric objects, etc
>- __aesthetics__: The word aesthetics refers to __the information__ in a plot. For example, which variables are associated with the x and y axes? We specify this using the `aes` function.
>- __geometric objects__: Geometric objects ("geoms") determine how the information is displayed. For example, will it be a scatter plot or a bar plot? We can specify geoms by adding a layer via functions beginning with `geom_`.

In practice, we build up a plot by combining different functions using the `+` operator. This has nothing to do with numeric addition!

## Illustrative example {.build}

<div>
Set up the basic object--define a default data frame and aesthetic...
```{r, eval=FALSE}
ggplot_object <- ggplot(my_df, aes(x = var1, y = var2))
```
</div>

<div>
Add a layer using the point 'geom'...
```{r, eval=FALSE}
ggplot_object <- ggplot_object + geom_point()
```
</div>

<div>
Show the plot--just 'print' the object to the console
```{r, eval=FALSE}
ggplot_object
```
</div>

## Real example: scatter plots 

Scatter plots are used to show the relationship between 2 continuous variables. Using the `iris` dataset, let's examine the relationship between petal length and petal width. 

__STEP 1__:

We use the `aes` function inside the `ggplot` function to specify which variables we plan to display. We also have to specify where the data are: 
```{r}
plt <- ggplot(iris, aes(x = Petal.Width, y = Petal.Length))
```
All we did here was make a `ggplot` object. 

## Real example: scatter plots
We can try to print the plot to the screen: 
```{r}
plt
```
This produces an empty plot because we haven't added a layer using a `geom_` function yet.

## Real example: scatter plots {.smaller}

__STEP 2__:

We want to make a scatter plot so we need to use the `geom_point` function:
```{r}
plt <- plt + geom_point()
```
Notice that all we do is "add" the required layer. Now we have something to plot:
```{r, eval=FALSE}
plt
```
```{r, fig.height=2.6, fig.width=3.2, echo=FALSE}
print(plt)
```

## Real example: scatter plots {.smaller}

__STEP 3__:

Maybe we should improve the axis labels? To do this, we need to "add" labels information using the `labs` function
```{r, eval=FALSE}
plt <- plt + labs(x = "Petal Width (cm)", y = "Petal Length (cm)")
plt
```
This just adds some new information about labelling to the prexisting `ggplot` object. Now it prints with improved axis labels:
```{r, fig.height=2.6, fig.width=3.2, echo=FALSE}
plt <- plt + labs(x = "Petal Width (cm)", y = "Petal Length (cm)")
plt
```

## Example: Scatter plots

__Doing it all in one go...__:

We don't have to build a plot object up in separate steps and then explicitly "print" it to the Console. If we just want to make the plot in one go we can do it like this:
```{r, fig.height=2.6, fig.width=3.2}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point() + 
  labs(x = "Petal Width (cm)", y = "Petal Length (cm)")
```

## Exercise

<div id="exercise-box">
__Customising your plot__
<div id="box-text">
Repeat the example we just stepped through, but now try to customise the point colours and their size. If that's too easy, see if you can make the points transparent. An example of suitable output is given below.

Hint: The `geom_point` function is responsible for altering these features (it takes a number of different arguments). It has arguments that control things like colour and size. They have sensible names...
</div>
</div>

```{r, fig.height=2.6, fig.width=3.2, echo=FALSE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Petal Width (cm)", y = "Petal Length (cm)")
```

## Answer

```{r, fig.height=2.6, fig.width=3.2, include=showAnswers}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Petal Width (cm)", y = "Petal Length (cm)")
```

## Adding more information {.build}

Q: The last graph was quite nice, but what information was missing?

A: The species identity!

```{r, fig.height=2.9, fig.width=4.7, echo=FALSE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) + 
  geom_point(size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

```{r, eval=FALSE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) + 
  geom_point(size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Exercise {.build}

<div id="exercise-box">
__Aesthetic mappings vs. arguments to `geom_`__
<div id="box-text">
You may have noticed that we can set something like "colour" in two places: the aesthetic mapping (`aes`) or via an argument to a geom (`geom_`). What happens if we set the colour in both places at once? Experiment with the iris petal length vs. petal width scatter plot example to work this out. Which one---the aesthetic mapping or geom argument---has precedence?
</div>
</div>

## Exercise {.build}
```{r, fig.height=2.4, fig.width=2.9, include=showAnswers}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Exercise: Putting it all together

We want to make the following scatter plot. It shows mean `wind` speed against mean `pressure`, where the means are calculated for each combination of storm `name` and `type`. The storm type of each point is delineated by its colour.

```{r, echo = FALSE, fig.width=6}
storms %>% 
  group_by(name, type) %>%
  summarise(wind = mean(wind), pressure = mean(pressure)) %>%
  ggplot(aes(x = pressure, y = wind, col = type)) + 
    geom_point(alpha = 0.7) + 
    labs(x = "Mean pressure (mbar)", y = "Mean wind speed (mph)")
```

## Exercise {.build}

<div id="exercise-box">
__Using `dplyr` and `ggplot2` together (part 1)__
<div id="box-text">
The first step is to work out how to use `dplyr` to calcuate the mean `wind` speed and mean `pressure` for each combination of storm `name` and `type`. Do this with the pipe (`%>%`) operator, and give the resulting data the name `storms.summary`.
</div>
</div>

```{r, include=showAnswers}
storms.summary <-
  storms %>% 
  group_by(name, type) %>%
  summarise(wind = mean(wind), pressure = mean(pressure))
```

## Exercise {.build}

<div id="exercise-box">
__Using `dplyr` and `ggplot2` together (part 2)__
<div id="box-text">
The next step uses the `storms.summary` data to plot the mean `wind` speed and mean `pressure` for each name-type case. Remember to colour the points by `type`.
</div>
</div>


## Exercise {.build}
```{r, fig.width=6, include=showAnswers}
ggplot(storms.summary, 
       aes(x = pressure, y = wind, col = type)) + 
  geom_point(alpha = 0.7) + 
  labs(x = "Mean pressure (mbar)", y = "Mean wind speed (mph)")
```

## Exercise {.build}

<div id="exercise-box">
__Using `dplyr` and `ggplot2` together (part 3)__
<div id="box-text">
Finally, see if you can combine the solutions to part 1 and 2 into a single "piped" operation. That is, instead of storing the intermediate data in `storms.summary`, use the pipe (`%>%`) to send the data straight to `ggplot`.
</div>
</div>

## Exercise {.build}
```{r, fig.width=6, include=showAnswers}
storms %>% 
  group_by(name, type) %>%
  summarise(wind = mean(wind), pressure = mean(pressure)) %>%
  ggplot(aes(x = pressure, y = wind, col = type)) + 
    geom_point(alpha = 0.7) + 
    labs(x = "Mean pressure (mbar)", y = "Mean wind speed (mph)")
```

Histograms | Visualising a single variable
====================================================================

## What are histograms? {.smaller}

Histograms summarise the relative frequency of different values of a variable. Look at the first 56 values of `pressure` variable in `storms`:

```{r}
storms $ pressure[1:56]
```

To get a sense of how frequent different values are we can "bin" the data. Here are the frequencies of `pressure` variable values, using 8 bins:

```{r}
table(cut(storms $ pressure, breaks = 8))
```

(You don't need to remember this R code)

## What are histograms? {.smaller}

We use histograms to understand the __distribution__ of a variable. They summarise the number of observations occuring in a contiguous series of bins. We can use `geom_histogram` to construct a histogram. Here is an example:

```{r}
ggplot(storms, aes(x = pressure)) + 
  geom_histogram(colour = "darkgrey", fill = "grey", binwidth=10) + 
  labs(x = "Pressure", y = "Count")  
```

## Exercise

<div id="exercise-box">
__Plotting histograms__
<div id="box-text">
Working with the `iris` dataset, construct a histogram of the ratio of petal length to petal width. See if you can make you histogram look like the one below.   
Hint: you can carry out the calculation with `Petal.Length` and `Petal.Width` inside `aes` (you don't have to first use `mutate` from `dplyr`)
</div>
</div>

```{r, echo=FALSE}
ggplot(iris, aes(x = Petal.Length / Petal.Width)) + 
  geom_histogram(binwidth=0.5) + 
  labs(x = "Petal Eccentricity", y = "Count")  
```

## Answer

```{r, include=showAnswers}
ggplot(iris, aes(x = Petal.Length / Petal.Width)) + 
  geom_histogram(binwidth=0.5) + 
  labs(x = "Petal Eccentricity", y = "Count")  
```

Alternative to histograms | visualising 'small' data
====================================================================

## Dot plots

We use dot plots to explore the __distribution__ of variables when we have relatively few observations (e.g. < 100). Here is an example:

```{r}
setosa <- filter(iris, Species == "setosa")
ggplot(setosa, aes(x = Sepal.Length)) + 
  geom_dotplot(binwidth=0.1)  
```

## Saving plots

We can save our plots using the `ggsave` function as follows

```{r}
ggplot(setosa, aes(x = Sepal.Length)) + 
  geom_dotplot(binwidth=0.1) + 
  ggsave("Sepal_dotplot.pdf", height = 4, width =4)
```

## Exercise
<div id="exercise-box">
__Saving plots__
<div id="box-text">
Working with the `iris` dataset, construct a dotplot of petal length for Iris virginica.  
Use ggsave to save this plot. Can you work out where R has saved you plot to? (i.e. which folder on your computer)
</div>
</div>

## Answer
```{r, include=showAnswers}
virginica <- filter(iris, Species== "virginica")
ggplot(virginica, aes(x = Sepal.Length)) + 
  geom_dotplot(binwidth=0.1) + 
  ggsave("Sepal_dotplot.pdf", height = 4, width =4)
```


## What happens next?

<br><br>

1. __No lecture__ tomorrow.

2. Course book, course book, course book! 

3. Let me know if you need help

4. More plotting next week (Then we're done...)
