---
layout: default
title: Exercises - functions and vectors 
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=7, fig.height=8, warning=TRUE, error=TRUE, prompt=FALSE)
```

Miscellaneous stuff (not used)
====================================================================

## Density plots

We use density plots to explore the __distribution__ of variables when we have a large number of observations (e.g. 1000s). Here is an example:

```{r}
dplot <- ggplot(storms, aes(x = pressure))
dplot <- dplot + geom_density() + labs(x = "Pressure", y = "Density")  
print(dplot)
```

## Density plots

We can also plot several different density plots at the same time using the `group` argument in `aes`. We can (optionally) colour these seperately using the `colour` argument in `aes`. Here is an example:

```{r, fig.width=5}
dplot <- ggplot(storms, aes(x = pressure, colour = type))
dplot <- dplot + geom_density() + labs(x = "Pressure", y = "Density")  
print(dplot)
```

## Exercise 

<div id="exercise-box">
__Plotting histograms__
<div id="box-text">
Working with the `diamonds` dataset, construct a single figure that shows density plots of `price` for each `cut` class. Use a log transformation of `price`. See if you can make your figure look like the one below.
</div>
</div>

```{r, echo=FALSE}
dplot <- ggplot(diamonds, aes(x = log10(price), group = cut, colour = cut))
dplot <- dplot + geom_density() + 
                 labs(x = "Price", y = "Density")  
print(dplot)
```

## Answer

```{r, eval=showAnswers, echo=showAnswers}
denplot <- ggplot(diamonds, aes(x = log10(price), group = cut, colour = cut))
denplot <- denplot + geom_density() + labs(x = "Price", y = "Density")  
print(denplot)
```


### Question XXXX

Take the natural log of 5 and round this to 3 decimal places. Do this with just one line of R code, i.e. make sure you do the whole calculation in one step.

#### Answer

The `log` function takes the natural log of a number (or more accurately, it takes the natural log of the elements of a numeric vector). This function only needs a single argument (called `x`), which is the numeric vector. The `round` function rounds the elements of a numeric vector to a user-specified number of decimal places. This takes two arguments: `x` is the numeric vector, and `digits` is an integer specifying the decimal places. In order to use these two functions together you have to remember to use function nesting, like this:

```{r}
round(x = log(5), digits = 3)
```

You could also have done this without naming either of the arguments...

```{r}
round(log(5), 3)
```

...though it is generally safer to name all but the first argument...

```{r}
round(log(5), digits = 3)
```
