---
title: "Exercises - functions and vectors"
output:
  html_document:
    theme: united
    highlight: zenburn
---

```{r, include=FALSE}

mk_link <-  function (title, page) {
  base_url <- "https://dzchilds.github.io/eda-for-bio/"
  paste0("[", title, "](", base_url, page, ".html)")
}

url_big_calc   <- mk_link("Using R as a big calculator",   "a-quick-introduction-to-r")
url_functions  <- mk_link("Using functions",               "using-functions")
url_num_vecs   <- mk_link("Numeric vectors",               "numeric-vectors")
url_other_vecs <- mk_link("Other kinds of vectors",        "other-kinds-of-vectors")
url_subsets    <- mk_link("Extracting subsets of vectors", "extracting-subsets-of-vectors")

i <- 0
```

<div class="well">
Today's practical provides an opportunity to practice carrying out some basic calculations with vectors. It requires knowledge of the material in the `r url_big_calc`, `r url_functions`, `r url_num_vecs`, `r url_other_vecs` and `r url_subsets` chapters. 

If you've already managed to work through those chapters and you're feeling comfortable with the material they contain the practical can be done in about 45 minutes. If not, you will need to spend some time reading the relevant material. The TAs can tell you where to look for help answering each question and will provide further assistance when you're stuck. There's no need to spend ages on a problem if you do get stuck. Ask for help if you aren't making progress after a few minutes of trying, and don't forget to help one another!

Put all of your working into a well-commented (lot's of `#`) and organised script. Once finished, you should be able to restart R and rerun your script without errors. Make sure you do this before leaving the practical. Answers will be provided on MOLE next week.
</div>


### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_big_calc`

The total volume (V) of a sphere is given by this equation:

$$
V = 4/3 \times \pi r^3
$$

Write some R code to calculate the volume of a sphere when the radius is:

a) 4
b) 5
c) 6

Now write some R code to calculate how much bigger the radius 5 sphere is compared to the radius 4 sphere.

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_functions` and `r url_num_vecs`

Make a vector of the first 100 integers ("whole numbers"). It should look like this when printed to the Console:

```{r, echo=FALSE}
1:100
```

Find the mean and standard deviation of the elements of this vector. Hopefully you know what the mean is. The standard deviation of a set of numbers is a measure of how variable they are. Ask a TA to explain it if you have not come across it before and want to know more. In either case, some Googling should lead you to the right functions...

**Hint:** Don't try to do the calculations by looking up the formulas for the mean and variance. You will need to find the names of the functions that calculate these quantities. 

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_num_vecs`

Use the `c` function to make a pair of numeric vectors containing the two following sets of numbers:

A) 7, 2, 5, 3, 1 
B) 1, 4, 8, 3, 9

Make sure you assign them the names `A` and `B`.

Write some R code to determine the result of `A + B` and `A + B + 6`. 

Look carefully at the results. What happens when we add two numeric vectors together? What happens when we add a single number to a vector? Do you see how "vectorised" operations work in R?

Now make a new numeric vectors containing the following set of numbers (call it `X`):

X) 1, 2

Add this to the `A` vector. What happens when you add numeric vectors with different lengths?

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_num_vecs`

Create a numeric vector called `my_nums` with four elements: 3, 5, 7, and 9. It should look like this when printed to the Console:
```{r, echo=FALSE}
c(3, 5, 7, 9)
```
Now create a second vector, called `multip`, that produces the following vector when multiplied by `my_nums`:
```{r, echo=FALSE}
c(3, 5, 7, 9) * c(1, 2, 3, 4)
```

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_num_vecs`

Construct a vector that contains an ascending sequence of numbers where first element of the vector is 0 and the last element is 101, with successive numbers increasing in steps of 0.01 units. Assign the vector a name so that it is stored in the Global Environment. Look at the Environment tab in RStudio. Can you see how long the vector is (i.e. find out how many elements the vector contains)? Now use R to calculate the length of this vector.

**Hint:** You can often guess the name of functions that carry out a particular task. Foe example, `seq` makes a sequence and `rep` makes repeated/replicated sequences. You can guess the name of the function that calculates the length of something in R...

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_num_vecs`

Make a repeated sequence of whole numbers ("integers") that has the following pattern:
```{r, echo=FALSE}
rep(1:3, each = 2, times = 4)
```
This is only a partial sequence. Your vector should contain this repeating sequence but have exactly 240 elements.

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_subsets`

The square bracket construct (`[`) is used to subset ("extract") the elements of a vector. For example, we can access one element of a vector by its position using the square bracket notation:

```{r}
# make a length-4 numeric vector
my_vec <- c(1.1, 2.3, 4.0, 5.7)
# now extract the second element
my_vec[2]
```

Imagine that you want to access the last element of a vector, but you don't know how long it is (i.e. you can;t use `my_vec[4]`). How would you do it? Use the `my_vec` vector above to illustrate the solution.

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_num_vecs` and `r url_subsets`

Copy the following into your script:

```{r}
num_seq <- rep(1:3, each = 5, times = 5)
```

Now prepare some R code to subset the `rep_seq` vector so that you keep every other element, starting at the element in position 2, i.e. subset `rep_seq` so that you end up with the second element, the fourth element, the sixth element, and so on.

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_num_vecs` and `r url_subsets`

Create a numeric vector containing a decreasing sequence of integers ("whole numbers") from 14 to 3. Assign this to a name of your choice, and then change the 6th element to have a value of 14. 

**Hint:** We can combine the subsetting operator `[` with assignment (` <- `) to change the individual elements of a vector.

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_other_vecs`

Create a vector of integers from 0 to 9 and assign this the name `x`. It should look like this when printed to the Console:

```{r, echo=FALSE}
0:9
```

Create an evenly spaced sequence of 10 numbers from 3 to 7 and assign this the name `y`. It should look like this when printed to the Console:

```{r, echo=FALSE}
seq(from = 3, to = 7, length = 10)
```

Use these two vectors to create a **logical vector** indicating which elements in `x` are larger than the corresponding elements in `y`. 

**Hint:** This requires you to know about  __relational operators__. These are discussed in the relevant chapter.

### Question `r {i <- i + 1; i}`

The pH of a solution is given by the negative of the logarithm (base 10) of the concentration of hydrogen ions, measured in moles per liter,  Calculate the pH of a river sample with a H+ concentration of 2.57 x 10^-7^ M l^-1^. Round your answer to 2 decimal places.

**Hint:** This is a simple calculation, but you first have to understand how R deals with very large and very small numbers. This means you have to understand [scientific notation](https://en.wikipedia.org/wiki/Scientific_notation) in R. It's impractical to write out all the zeros when writing a small number like 0.0000012, so instead we usually write such numbers as 1.2 x 10^-6^ in text. R allows us to represent such a number as `1.2e-06`---the "e-06" here stands for "x 10^-6^", i.e. it moves the decimal 7 places to the left.

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_functions` and `r url_num_vecs`

Use the `seq` function to create a vector containing a sequence of numbers between -1 and 1 in steps of 0.25. It should look like this when printed to the screen:

```{r, echo=FALSE}
seq(from = -1, to = +1, by = 0.25)
```

Use this vector (as the `x` argument) with the function `sinpi`. What is returned?

Now make a new vector of length 9, containing a sequence of numbers that starts at -$pi$ and ends at $pi$ (use the built-in constant `pi`).  It should look like this when printed to the screen:

```{r, echo=FALSE}
seq(from = -pi, to = +pi, length.out =  9)
```

Use this new vector (as the `x` argument) with the `sin` function. 

Do you understand the difference between the `sinpi` and `sin` functions? Are the answers exactly equal? 

**Hint:** Remember that R uses [radians](https://en.wikipedia.org/wiki/Radian) to define angles. Mathematicians don't measure angles in degrees. Instead, they measure angles in units called radians. There are 360 degrees in one rotation of a circle, but one full rotation of a circle corresponds to $2\pi$ radians. 

### Question `r {i <- i + 1; i}`

Relevant chapters: `r url_other_vecs` and `r url_subsets`

There are a number of different vectors built into R, each of which can be accessed even though they don't appear in the Global Environment. For example, the `LETTERS` and `letters` objects are character vectors containing the letters of the alphabet in upper and lower case characters, respectively:
```{r}
LETTERS
letters
```

Using the built-in `LETTERS` vector, make a character vector with elements that spells your first name in capital letters. Print this to the Console. For example, if your name is "Dylan", you should see:
```{r, echo=FALSE}
LETTERS[c(4, 25, 12, 1, 14)]
```

**Hint:** This question uses ideas from the `r url_other_vecs` and `r url_subsets` chapters.

### Question `r {i <- i + 1; i}`

Create a vector containing the letters of the alphabet in reverse order. 

**Hint:** use the `:` operator or the `seq` function to extract elements of the built-in vector `LETTERS` (or `letters`, either is fine).

### Question `r {i <- i + 1; i}` (Optional extended exercise)

*Brief Background:* The unified neutral theory of biodiversity is a hypothesis that an individual’s species identity is not related to its survival or fitness.  Under neutral theory, mortality and recruitment are random and population size is only constrained by a maximum community size.  While we know that species traits are important at small spatial scales, neutral theory is interesting because it describes many of the patterns we see at larger scales. 

Here you will simulate a very simple tree community under neutral theory. The basic idea is that we label tree "species" and then let them "reproduce" at random (i.e. we imagine they are all the same). Here are the steps to build a very basic simulation of a neutral community:

a)	Create a character vector containing the letters “a” through “j” and assign it to the name `comm`. This will represent the community, with each element representing an individual and each letter representing a species. **Hint:** use the inbuilt `letters` vector.

b)	Calculate the species richness of this community. Species richness is just the number of unique species in the community. **Hint:** use functions `unique` and `length`.

c)	This bit is quite tricky. We need to be able to choose a random individual to successfully reproduce at each time step. Work out how to select out a random element of `comm` (using R of course!). **Hint:** Use the `sample` function, working with the `x` and `size` arguments. 

d)	We also need to select a random individual to die and be replaced at each time step. To do this, pick out one number at random from the sequence generated by `1:10` (use the same trick you learned above). This will give you the index (i.e. position) of the element to be replaced.

e)	Use square bracket (`[]`) indexing and your answer to (d) to pick out the letter of the individual in `comm` to be replaced.

f)	Combine your answers to (c) and (e) to replace a random element with another. **Hint:** use assignment---the thing on the left to be replaced by the thing on the right.

g)	Run the last step (f) a further 9 times, what has happened to the species richness? What would happen to this community given enough time steps?
