---
title: Packages, help and data frames
author: Dylan Childs
date: February 23, 2017
highlighter: pygments
output: 
  ioslides_presentation:
    css: extras.css
    smaller: false
---

```{r, echo=FALSE}
showAnswers <- TRUE
```

One last bit of the basics... | Logical comparisons
====================================================================

## Comparing vectors {.build .smaller}

<div>
We use __relational operators__ to compare the elements of numeric vectors:

- `x < y `: is x less than y?
- `x > y `: is x greater than y?
- `x <= y`: is x less than or equal to y?
- `x >= y`: is x greater than or equal to y?
- `x == y`: is x equal to y?
- `x != y`: is x not equal to y?

These operators produce something called a __logical vector__ (a vector of TRUE and FALSE values---nothing else allowed)
</div>

<div>
__Example__
```{r}
x <- seq(1, 9)
y <- rep(5, 9)
x == y
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Using relational operators__
<div id="box-text">
Make two vectors named `num1` and `num2`, where `num1` is a sequence from 0 to 20 in steps of 2, and `num2` is a sequence from 5 to 15 in steps of 1. You can do it like this: `num1 <- seq(0, 20, 2)` and `num2 <- seq(5, 15, 1)`. Once you have made these print them to screen so that you know exactly what they look like, then experiment with `<`, `>`, `<=`, `>=`, `==` and `!=` to convince yourself you understand how the work.
</div>
</div>

```{r, echo=FALSE}
num1 <- seq(0, 20, 2)
num2 <- seq(5, 15, 1)
```

<div>
```{r, include=showAnswers}
num1
num2

```
</div>

## Exercise {.build}

<div>
```{r, include=showAnswers}
num1
num2
```
</div>

```{r, include=showAnswers}
num1 == num2
num1 != num2
```

## Exercise {.build}

<div>
```{r, include=showAnswers}
num1
num2
```
</div>

```{r, include=showAnswers}
num1 <  num2
num1 <= num2
```

## Combining comparisons {.build .smaller}

<div>
We use __logical operators__ to combine logical statements:

- `!x `: __negation__ = `TRUE` becomes `FALSE`, `FALSE` becomes `TRUE`
- `x & y `: __AND__  = `TRUE` if __both__ x and y are `TRUE`, `FALSE` otherwise
- `x | y`: __OR__  = `TRUE` if __either__ x and y are `TRUE`, `FALSE` otherwise

These operators produce another __logical vector__ (a vector of TRUE / FALSE values)
</div>

<div>
__Example__
```{r}
x <- c(TRUE, TRUE, FALSE, FALSE)
y <- c(FALSE, TRUE, FALSE, TRUE)
x & y
x | y
```
</div>

The working directory | Where R "looks" for files
======================================================================================

## What is the working directory? {.build}

A quick overview of what it is and how to work with it...

- The **working directory** is a default location where R looks for files you want to use. It is just a folder on your computer.

- If you don't set the working directory, R will do it for you. It will seldom choose the exact folder you want to use.

- YOU should set the working directory EVERY time you start R/RStudio. Setting your working directory should be the first thing you do.

- Don't do this using R code in your script. Instead, do it via the Console (hard and error prone) or using the RStudio menu system (easy and reliable).

<div>
**Set your working directory now using...** 
**`Session > Set Working Directory > Choose Directory...`**
</div>

Data frames | A place for your data
======================================================================================

## Two views of data frames {.build}

The first view: __Data frames are a collection of vectors__. These may be different types (numeric, character, others), but they must all be the same length.

<div>
Here is how to make one "by hand" using the `data.frame` function:

```{r}
mydf <- data.frame(Var1 = 1:4, Var2 = LETTERS[1:4], Var3 = c(1,2,4,8))
mydf
```

Data frames are central to data analysis in R. You can think of each vector as corresponding to a __statistical variable__ that we want to analyse (more on that in the course book).
</div>

## Selecting vectors from data frames {.build}

We can extract a vector from a data frame in a few different ways:

<div>
- Using the name and the `$` operator
```{r}
mydf $ Var1
```
- Using the name and the `[[ ]]` operator
- Using the position and the `[[ ]]` operator
</div>

## Selecting vectors from data frames

We can extract a vector from a data frame in a few different ways:

- Using the name and the `$` operator
- Using the name and the `[[ ]]` operator
```{r}
mydf[["Var1"]] # <- WE HAVE TO USE QUOTES!
```
- Using the position and the `[[ ]]` operator


## Selecting vectors from data frames

We can extract a vector from a data frame in a few different ways:

- Using the name and the `$` operator
- Using the name and the `[[ ]]` operator
- Using the position and the `[[ ]]` operator
```{r}
mydf[[1]]
```

<br><br><br><hr>

N.B. We will only use the first version in this course. Remember what `$` does. You **will** need to use it.

## Exercise {.build}

<div id="exercise-box">
__Extracting vectors from data frames__
<div id="box-text">
R has many datasets built into it, stored as data frames. One of these is the famous `iris` data. Type this into your Console: `View(iris)`. How many vectors (i.e. columns) are stored in `iris`? Extract the one named `Sepal.Width`.
</div>
</div>

<div>
```{r, include=showAnswers}
iris$Sepal.Width
```
</div>

## Two views of data frames {.build}

The second view: __Data frames are table-like objects__. Take another look at our simple data frame:

```{r}
mydf
```

This clearly has rows and columns. We can extract elements of the data frame using the square brackets `[ ]`:

```{r}
mydf[2, 3]
```

__[Row #, Col #]__: This extracted the element in the second row and third column. 

## Exercise {.build}

<div id="exercise-box">
__Extracting elements from data frames__
<div id="box-text">
See if you can use the `[,]` construct to get the elements in rows 2 to 3 __and__ columns 1 to 4 from the `iris` data. Hint: you will need to use the `:` operator, twice inside `[,]` 
</div>
</div>

<div>
```{r, include=showAnswers}
iris[2:3, 1:4]
```
</div>

__Note__: this produced a new data frame.

## Exercise {.build .smaller}

<div id="exercise-box">
__Examining a data frame__
<div id="box-text">
We've seen that `View(iris)` can be used to examine a data frame in a spreadsheet like view. There are many other functions that allow us to examine the structure of a data frame. Using the `iris` data, experiment with the `head`, `tail`, and `str` functions to see what they do.
</div>
</div>

<div>
```{r, include=showAnswers}
head(iris, n = 2)
```
</div>

<div>
```{r, include=showAnswers}
str(iris)
```
</div>


Reading in data | A great source of confusion
====================================================================

## We need some data....

**1. Grab the "iris_morphology.xlsx" file from MOLE**:

(APS135 Skills for Biologists > Information Technology  > Data files needed in semester 2)

**2. Save this anywhere you like**

Make sure you remember where you put it!

**3. Open up your copy of "iris_morphology.xlsx" in Excel**

How many columns ("variables") and rows ("observations") are there?

**4. Once you understand the data, close Excel down**

We're ready...

## Getting data from Excel into R {.build .smaller}

Essentially a two step process

<div>
**STEP 1. Export data from Excel to a CSV file**:

- Open the "iris_morphology.xlsx" spreadsheet up in Excel

- In Excel, go to `Save As...` and choose the `Comma Separated Values (csv)` option

- Keep the default file name ("iris_morphology.csv"), but make sure the file is saved **in your working directory** (you set this earlier)
</div>
<div>
__STEP 2. Now import the CSV file into using RStudio:__

- Include the following line in your script, and send it to the Console

```{r, eval = FALSE}
my.iris <- read.csv("iris_morphology.csv", stringsAsFactors = FALSE)
my.iris
```

- What happened? Did you see any error messages? If you did, check your spelling and make sure your working directory is set to the right place.
</div>

## The other way...

__This is how to import data in a CSV file using RStudio:__

- Select `Import Dataset > From Text File...`

- Select the file your saved and click Open

- Enter a name (no spaces allowed) and click Import

**Some advice: Don't use this method!!!** You have to remember do it everytime you want to run your script. This is very error prone.

R packages | The other great source of confusion 
====================================================================

## A (very) quick overview {.build}

<div>
Packages extend what base R can do. Two things have to happen before you can use a package:

1. You have to **install** it. This is a "do once" operation. After a package is installed you can close down and reopen R and it is still installed.

2. You have to **load and attach** it. This is a "do every time" operation. You have to load and attach the package every time you start a new session (= open R).
</div>

<hr>

<div>
A mobile phone App analogy: 

- installing a package <=> getting an App from an App store

- loading and attaching a package <=> opening the App on your phone
</div>

## How do we access packages?

1. We **install** a package with the `install.packages` function

2. We **load and attach** an installed package with the `library` function

**TIPS**

- Do not use `install.packages` inside your scripts! You don't want to reinstall a package every time you plan to use it. 

- You can also install a package via the Packages tab in RStudio. It is fine to use this if you don't like using `install.packages`.

## Exercise {.smaller}

<div id="exercise-box">
__Working with packages__
<div id="box-text">
We want to use the `fortune` function (singular!) that resides in the **fortunes** package (plural!). Do the following, in the order given, working directly in the Console:

1. Try to use the `fortune` function, i.e. type `fortune()` straight into the Console. Does this work or do you see an error?

2. Try to use `library` to load and attach the **fortunes** package. Does this work or do you see an error?

3. Look in the **Packages** tab in RStudio. Can you see anything called fortunes? Now install the **fortunes** package with `install.packages`. Look for it the **Packages** tab again. 

4. Try again to use the `fortune` function. Does this work? If not, why not?

5. Use `library` to load and attach the **fortunes** package (plural!). What happens?

6. Try again to use the `fortune` function. Does this work or do you see an error?
</div>
</div>

Getting help | Learning about a function
======================================================================================

## Help files

Do the following:

```{r, eval=FALSE}
?round
```

Help files are really **documentation** files:

- Description
- Usage
- Arguments
- Details
- References
- Examples

Let's step though the help for `round`...

## Other sources of help {.smaller .build}

The internet is packed full of helpful resources:

1. **Google** -- this might be obvious, but you have to work on your "Google-fu" to find what you want. The key is to pick the right search terms:

    - include the function or package name
    
    - include the letter R!
    
    - be selective about which sites you look at...

<div>
2. **Stack Overflow** -- probably the best question and answer site for computing stuff. People can be a bit blunt though:

    - check whether your question has been asked before (Google)
    
    - read the community guidelines
    
    - provide example code with your questions
</div>

## What happens next?

<br><br>

1. __No lecture__ tomorrow.

2. Keep reading through the online course book.

3. Reminder: DZC office hours are 12-3pm every Tuesday.

4. We'll start getting the useful stuff next week. I promise!




