---
title: Data wrangling with dplyr
author: Dylan Childs
date: March 1, 2017
highlighter: pygments
output: 
  ioslides_presentation:
    css: extras.css
    widescreen: true
    smaller: true
---

```{r, include=FALSE}
showAnswers <- TRUE
library(dplyr)
library(nasaweather)
```

Overview of **dplyr** and `tbl` | Getting to grips with the basics 
====================================================================

## Why use dplyr?

**dplyr** implements a __grammar of data manipulation__ to enable you manipulate data and summarise the information in a data set (e.g. group means).

__Advantages of using **dplyr**__
 
- Provides a consistent and easy to understand framework for data manipulation
- Designed to work well with the `ggplot2` plotting system
- Fast compared to many base R functions 
- Allows you to work with data stored in many ways (e.g. in a database)

## Five key verbs {.build} 

<div>
__**dplyr** has five main "verbs" (i.e. functions):__

- `select`: Extract a subset of variables
- `filter`: Extract a subset of rows
- `arrange`: Reorder rows 
- `mutate`: Construct new variables 
- `summarize`: Calculate information about groups

(+ a few other important functions, e.g. `slice`, `rename`, `transmute`, and `group_by`)
</div>

<div>
<br>
__It is helpful to classify the verbs according to what they work on:__

- observations (rows): `filter` & `slice` & `arrange`
- variables (columns): `select` & `rename` & `mutate`
- summarise: `summarize` (or `summarise`)

(This classification only works if your data are __tidy__. Make sure you read about this)
</div>

## Before we start {.build}

Required packages and data:

- **dplyr**: obviously
- `storms.csv`: climate data for Central America (on MOLE)

<div>
**1. Load the __dplyr__ package**

```{r, message=FALSE}
library(dplyr)
```

This should already be installed. If you see errors, you may have to install it, e.g.

```{r, eval=FALSE}
install.packages("dplyr")
```
</div>

<div>
__2. Read in the storms data__

Download the `storms.csv` data from MOLE. Set your working directory, and then use `read.csv` to read the data in (remember to use `stringsAsFactors = FALSE`). Assign the resulting data frame the name `storms`.
</div>


__3. Use `View` to examine the `storms` data__

## Making and using `tbl` objects

A `tbl` object (pronounced "tibble") is essentially a special kind of data frame. It works the same as a data frame, but with a few small differences... e.g. compact printing:

```{r}
tbl_df(storms)
```


## Convert `storms` and `iris` to tibbles now

Next in your script, add these lines tConvert `storms` and `iris` to tibbles...

```{r}
storms <- tbl_df(storms)
iris   <- tbl_df(iris)
```

You don't have to do this (**dplyr** is fine with normnal data frames), but it will ensure your output matches the presentation.

## Looking at your `tbl` (or `data.frame`) 

In addition to printing a `tbl` or `data.frame`, we can use the `glimpse` function with these objects to obtain different summary information about variables:

```{r}
glimpse(storms)
```

Glimpse tells us what variables are in `storm` as well as the type of each variable.

Using `select` | Extracting a subset of variables 
====================================================================

## Basic usage

We use `select` to to extract a subset of variables for further analysis. Using `select` looks like this:

```{r, eval=FALSE}
select(data, Variable1, Variable2, ...)
```

__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `VariableX`: names of variables in `data`

## Exercise {.build}

<div id="exercise-box">
__Selecting two variables__
<div id="box-text">
Use the `select` function with the `storms` data set to make a new data set containing only `name` and `year`. Assign this new data set a name, and then check that it contains the right variables using the `glimpse` function.
</div>
</div>

```{r, include=showAnswers}
storms_simple <- select(storms, name, year)
glimpse(storms_simple)
```

## Selecting & dropping variable ranges {.build}

The `select` function makes selecting/removing groups of variables easy:

- Use `:` to select a sequence of variables 
- Use `-` to drop a sequence of variables 

The sequence can be specified using numbers (for position) or names.

<div>
__Usage:__

```{r, eval=FALSE}
# a range of variable to keep
select(data, Variable1:Variable5)
# a range of variable to drop
select(data, -(Variable1:Variable5))
```
</div>

<div>
__Example:__

```{r}
iris_fewer <- select(iris, Petal.Length:Species)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Selecting a range of variables__
<div id="box-text">
Use the `select` function with the `storms` data set to select __just__ the variables `name`, `year` and `month` variables.
</div>
</div>

```{r, include=showAnswers}
storms_fewer <- select(storms, name:month)
glimpse(storms_fewer)
```

```{r, include=showAnswers}
# alternatively
x <- select(storms, -(day:seasday))
```

## Helper functions {.build}

There are several helper functions that work with `select` to simplify common variable selection tasks:

- `starts_with("xyz")`: every name that starts with `"xyz"`
- `ends_with("xyz")`: every name that ends with `"xyz"`
- `contains("xyz")`: every name that contains `"xyz"`
- `matches("xyz")`: every name that matches `"xyz"`
- `one_of(names)`: every name that appears in `names` (character vector).

<div>
__Usage:__

```{r, eval=FALSE}
select(data, help_func("xyz"))
```
</div>

<div>
__Example:__

```{r, eval=FALSE}
iris_petal <- select(iris, starts_with("Petal"))
glimpse(iris_petal)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Using select with helper functions__
<div id="box-text">
Use the `select` function with the `storms` data set to create a new data set containing just the `lat` and `long` variables. Do this using the `starts_with` helper function inside select. Then see if you can work out how to select just the `name` and `type` variables using the `one_of` helper function.
</div>
</div>

```{r, include=showAnswers}
storms_fewer <- select(storms, starts_with("l"))
glimpse(storms_fewer)
```

```{r, include=showAnswers}
storms_fewer <- select(storms, one_of("name", "type"))
glimpse(storms_fewer)
```


Using `select` and `rename` | Renaming variables
====================================================================

## Renaming while selecting {.build}

We can use `select` to rename variables as we select them using the `newName = varName` construct.

<div>
__Usage:__

```{r, eval=FALSE}
select(data, newName1 = Var1, newName2 = Var2, ...)
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_select <- select(iris, PetalLength = Petal.Length)
glimpse(iris_select)
```
</div>

## Renaming without selecting {.build}

Use `rename` to rename variables while keeping all variables using the `newName = varName` construct.

<div>
__Usage:__

```{r, eval=FALSE}
rename(data, newName1 = Var1, newName2 = Var2, ...)
```
</div>

<div>
__Example:__

```{r}
iris_renamed <- rename(iris, 
                       PetalLength = Petal.Length, 
                       PetalWidth  = Petal.Width)
glimpse(iris_renamed)
```
</div>


Using `mutate` | Making new variables 
====================================================================

## Basic usage {.build}

<div>
We use `mutate` to __add variables__ for further analysis. This is how we use `mutate`:

```{r, eval=FALSE}
mutate(data, NewVar = <expression>, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `NewVar`: name of a new variable to create
- `<expression>`: an R expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` which appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`.

- You may use more than one `NewVar = <expression>` at a time if you need to construct several new variables.
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
Use the `mutate` function with the `iris` data set to make a new variable which is the petal area, $(Area = Length \times Width)$.
</div>
</div>

```{r, include=showAnswers}
iris_area <- mutate(iris, Petal.Area = Petal.Length * Petal.Width)
glimpse(iris_area)
```

## Multiple calculations {.build}

<div>
We can add more than one variable at a time using `mutate`. Each new variable can also use one or more variables created in a previous step.
</div>

<div>
__Usage:__

```{r, eval=FALSE}
mutate(data, NewVar1 = <expression1>, 
             NewVar2 = <expression2 using NewVar1>)
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_new_vars <- 
  mutate(iris, Sepal.Eccentricity = Sepal.Length / Sepal.Width,
               Petal.Eccentricity = Petal.Length / Petal.Width,
               Eccentricity.Diff  = Sepal.Eccentricity - Petal.Eccentricity)
glimpse(iris_new_vars)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Making several new variables__
<div id="box-text">
Use the `mutate` function with the `iris` data set to make two new area variables, one for petal and one for sepal. Create a third variable which is the ratio of the petal and sepal areas. Do all of this in one call to `mutate`, i.e. use `mutate` only once to do all of this.
</div>
</div>

```{r, include=showAnswers}
iris_ratio <- mutate(iris, Sepal.Area = Sepal.Length * Sepal.Width,
                           Petal.Area = Petal.Length * Petal.Width,
                           PS.Area.Ratio = Petal.Area / Sepal.Area)
glimpse(iris_ratio)
```

Using `filter` | Selecting subsets of observations 
====================================================================

## Basic usage {.build}

<div>
We use `filter` to select a subset of rows for further analysis, based on the result(s) of one or more logical comparisons. Using `filter` looks like this:

```{r, eval=FALSE}
filter(data, <expression>)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `<expression>`: an R expression that implements a logical comparison using variables in `data`
</div>

<div>
__Comments__

- The `<expression>` can be any valid R expression that uses variables in `data` and returns a logical vector of `TRUE` / `FALSE` values.
- The `<expression>` typically uses a combination of __relational__ (e.g. `<` and `==`) and __logical__ (e.g. `&` and `|`) operators
</div>

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on one variable__
<div id="box-text">
Use the `filter` function with the `storms` data set to create a new data set containing just the observations associated with storms classified as Hurricanes.

Hint: use `glimpse` to remind yourself of the variable names in `storms`. You need to work out which one contains information about the storm category.
</div>
</div>

```{r, include=showAnswers}
filter(storms, type == "Hurricane")
```

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on more than one variable__
<div id="box-text">
Repeat the last exercise, but now extract the observations associated with Hurricanes that took place in 1997 or later.
</div>
</div>

```{r, include=showAnswers}
filter(storms, 
       type == "Hurricane" & year >= 1997) 

# or use: filter(storms, type == "Hurricane", year >= 1997)
```


Using `arrange` | Reordering observations 
====================================================================

## Basic usage {.build}

<div>
We use `arrange` to reorder the rows of our data set. This can help us see associations among our variables. Using `arrange` looks like this:

```{r, eval=FALSE}
arrange(data, Variable1, Variable2, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `VariableX`: names of variables in `data`
</div>

<div>
__Comments__

- The order of sorting corresponds to the order that variables appear in `arrange`, meaning `data` is sorted according to `Variable1`, then `Variable2`, then `Variable3`, etc
- The sort order is from smallest to largest (ascending). If you want to reverse the sort order to go from largest to smallest (descending) use `desc(VariableX)`.
</div>

## Exercise {.build}

<div id="exercise-box">
__Reording observations__
<div id="box-text">
Use the `arrange` function to reorder the observations in the `storms` data set, according to the pressure variable. Store the resulting data set and use then use the `View` function to examine it. What can you say about the association between atmospheric pressure and storm category?
</div>
</div>

```{r, include=showAnswers}
storm.sort <- arrange(storms, pressure)
```

```{r, eval=FALSE}
View(storm.sort)
```

## What happens next?

<br><br>

1. __No lecture__ tomorrow (again!).

2. Keep reading the online course book (more chapters coming)!

3. Reminder (again): DZC office hours are 12-3pm every Tuesday.

4. We'll start learning to make plots next week













