---
title: More dplyr and ggplot introduction
author: Dylan Childs
date: March 10, 2016
highlighter: pygments
output: 
  ioslides_presentation:
    css: extras.css
    smaller: true
---

```{r, include = FALSE}
showAnswers <- TRUE
library(dplyr)
library(ggplot2)
library(nasaweather)
library(grid)
library(knitr)

opts_chunk$set(fig.width=4.5, fig.height=3.3, warning=FALSE, message=FALSE, 
               dev.args = list(bg = 'transparent'), fig.align='center')

my_theme <- theme_gray() + theme(
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(vjust=-0.5),
  axis.title.y = element_text(vjust=+1.5)
  )
theme_set(my_theme)
```

## Before we start {.build}

Required packages and data:

- `dplyr`: obviously
- `ggplot2`: new package
- `storms.csv`: climate data for Central America (on MOLE)

**1. Load the `ggplot2` package**

```{r, message=FALSE}
library("ggplot2")
```

This should already be installed. If you see errors, you may have to install it, e.g.

```{r, eval=FALSE}
install.packages("ggplot2")
```

Do the same for `dplyr` (you should NOT need to install it first)

__2. Read in the storms data__

Download the `storms.csv` data from MOLE. Set your working directory to the correct location and then use `read.csv` to read the data in, giving it the name `storms`.

A bit more `dplyr`... 
====================================================================

Using `summarise` | Calculating summaries of variables 
====================================================================

## Basic usage {.build}

<div>
We use summarise to __calculate summary statistics__ for further analysis. This is how we use `summarise`:

```{r, eval=FALSE}
summarise(data, SummaryVar = <expression>, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `SummaryVar`: name of your summary variable
- `<expression>`: an R expression that references variables in `data`
</div>

<div>
__Comments__

- The `<expression>` which appears on the right hand side of the `=` can be any valid R expression that uses variables in `data`. However, `<expression>` should return __a single value__.
- Although `summarise` looks a little like `mutate`, it is designed to construct a completely new dataset containing summaries of one or more variables.
- You may use more than one `SummaryStat = <expression>` at a time if you need to construct several summaries.
</div>

## Exercise {.build}

<div id="exercise-box">
__Calculating the mean of two variables__
<div id="box-text">
Use the `summarise` function with the `iris` dataset to calculate the mean sepal length and the mean sepal width.

Hint: You need to work out which R function calculates a mean. The clue is in the name.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
summarise(iris, mean.sl = mean(Sepal.Length), mean.sw = mean(Sepal.Width))
```

## Exercise {.build}

<div id="exercise-box">
__Calculating a more complex summary of a variable__
<div id="box-text">
Use the `summarise` function with the `iris` dataset to calculate the mean __area__ of sepals.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
summarise(iris, mean.sl = mean(Sepal.Length * Sepal.Width))
summarise(iris, mean.sl = mean(Sepal.Length) * mean(Sepal.Width))
```

Which one is right?

Using `group_by` | Making summaries for groups of observations
====================================================================

## Basic usage{.build}

<div>
We use `group_by` to add grouping information to a `tbl`. This is how we use `group_by`:

```{r, eval=FALSE}
group_by(data, GroupVar1, GroupVar2, ...)
```
</div>

<div>
__Arguments__

- `data`: a `data.frame` or `tbl` object 
- `GroupVar`: name of grouping variable(s) 
</div>

<div>
__Comments__

- The `group_by` does not do anything other than add grouping information to a `tbl`. It is only useful when used with `summarise` or `mutate`.
- Using `group_by` with `summarise` enables us to calculate numerical summaries on a _per group_ basis.
- Using `group_by` with `mutate` enables us to add variables on a _per group_ basis.
</div>

## Exercise {.build}

<div id="exercise-box">
__Using `group_by` to calculate group-specific means__
<div id="box-text">
Use the `group_by` function and the `summarise` functions with the `storms` dataset to calculate the mean wind speed associated with each storm __type__.

Hint: This is a two step exercise. Use `group_by` first to add some information to storms, remembering to assign the result a name. Then use `summarise` on this new dataset.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
storms.grp <- group_by(storms, type)
summarise(storms.grp, mean.wind = mean(wind))
```

## Exercise {.build}

<div id="exercise-box">
__Using `group_by` to group by more than one variable__
<div id="box-text">
Use the `group_by` and `summarise` functions with the `storms` dataset to calculate the mean and maximum wind speed (in the `wind` variable), associated with **each combination** of month and year. Assign the result a name and then use `View` to examine it. Which month in which year saw the largest maximum wind speed? 

Hint: You can guess the names of the two functions that calculate the mean and max from a numeric vector.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
storms.grouped <- group_by(storms, year, month)
summarise(storms.grouped, mean.speed = mean(wind), max.speed = max(wind))
```

## Exercise {.build}

<div id="exercise-box">
__Using `group_by` and `mutate`__
<div id="box-text">
Use the `group_by` and `mutate` functions with the `iris` dataset to calculate a "centred" version of sepal length. A centred variable is just one that has had its overall mean subtracted from every value. 

Do you understand the different behaviour of `summarise` and `mutate` when used alongside `group_by`?
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
iris.grp <- group_by(iris, Species)
mutate(iris.grp, SL.cen = Sepal.Length - mean(Sepal.Length))
```

Using `%>%` | piping or chaining
====================================================================

## Motivating example{.build}

It is common to need to perform a sequence of calculations on a vector or data frame, by applying a series of function in turn. We can do this in two ways:

<div>
**1. Method 1: Store intermediate results...** 


```{r}
x <- 10
x <- sqrt(x)
x <- exp(x)
round(x, 2)
```
</div>

<div>
**1. Method 2: Use function nesting...** 

```{r}
round(exp(sqrt(10)), 2)
```
</div>

These do the same thing. Method 1 is easy to read, but is very verbose. Method 2 is concise, but not at all easy to read.

## A third way, using `%>%`...{.build}

The `dplyr` package includes a special operator, called "the pipe". It looks like this: `%>%`. This allows us to avoid strong intermediate results (method 1), while still reading a sequence of function calls from left to right. For example:

```{r}
10 %>% sqrt(.) %>% exp(.) %>% round(., 2)
```

<div>
Or equivalently, and even simpler...

```{r}
10 %>% sqrt() %>% exp() %>% round(2)
```
</div>

## Why is this handy?{.build}

We can use `%>%` with any function we like. Look at this example of a two-step `group_by` and `summarise` operation with `iris`...

<div>
```{r, eval=FALSE}
# method 1
iris.grouped <- group_by(iris, Species)
summarise(iris.grouped, mean.SL = mean(Sepal.Length))
# method 2
summarise(group_by(iris, Species), mean.SL = mean(Sepal.Length))
```
</div>

<div>
The "piped" equivalent is more natural to read...

```{r, eval=showAnswers, echo=showAnswers}
iris %>% 
  group_by(Species) %>% 
  summarise(mean.SL = mean(Sepal.Length))
```
</div>


## Exercise {.build}

<div id="exercise-box">
__Using `group_by` to group by more than one variable__
<div id="box-text">
Let's repeat the earlier `group_by` and `summarise` exercise, using the pipe. Rememeber, we used the `group_by` and `summarise` functions with the `storms` dataset to calculate the mean and maximum wind speed (in the `wind` variable) associated with each combination of month and year.
</div>
</div>

```{r, eval=showAnswers, echo=showAnswers}
storms %>% 
  group_by(year, month) %>%
  summarise(mean.speed = mean(wind), max.speed = max(wind))
```

Moving on to `ggplot2`...
====================================================================

Overview of `ggplot2` | The grammar of graphics
====================================================================

## Why use `ggplot2`? {.build}

<div>
Roughly speaking, there are three commonly used plotting frameworks in R. 

- base graphics: available as soon as you open R
- `lattice` package: create __trellis__ graphs
- `ggplot2` implements a __grammar of graphics__.
</div>

<div>
__Advantages__ of using `ggplot2`
 
- Consistent and intuitive framework for plotting
- Fleixble enough to make every plot you will need
- Works well with `dplyr`
</div>

<div>
__Disadvantages__ of using `ggplot2`

- You have to learn "the grammar" to use it well
- Vast package, can be intimidating
- More than one way to do things
</div>

## Key concepts {.build}

You need to wrap your head around a few ideas to start using `ggplot2` effectively:

>- __layers__: We build ggplot2 objects by adding one or more layers together. We can do this in a stepwise way, only plotting the object when we are ready. Each layer needs things like data, aesthetics, geometric objects, etc
>- __aesthetics__: The word aesthetics refers to __the information__ in a plot. For example, which variables are associated with the x and y axes? We specify this using the `aes` function.
>- __geometric objects__: Geometric objects ("geoms") determine how the information is displayed. For example, will it be a scatter plot or a bar plot? We can specify geoms by adding a layer via functions beginning with `geom_`.

In practise, we build up a plot by combining different functions using the `+` operator. This has nothing to do with numeric addition!

<div>
__Illustrative example__

```{r, eval=FALSE}
# 1. define a default data frame and aesthetic
ggplot.object <- ggplot(my_df, aes(x = var1, y = var2))
# 2. add a layer with the point geom
ggplot.object <- ggplot.object + geom_point()
# 3. show the plot
print(ggplot.object)
```
</div>

## Example: Scatter plots

Scatter plots are used to show the relationship between 2 continuous variables. Using the `iris` dataset, let's examine the relationship between petal length and petal width. 

__STEP 1__:

We use the `aes` function inside the `ggplot` function to specify which variables we plan to display. We also have to specify where the data are: 
```{r}
plt <- ggplot(iris, aes(x = Petal.Width, y = Petal.Length))
```
All we did here was make a `ggplot` object. We can try to make the plot using the `print` function: 
```{r, eval=FALSE}
print(plt)
```
This fails, because we haven't added a layer using a `geom_` function yet. 

(N.B. In version 2.0 and above of `ggplot2`, using `print` on an incomplete object will plot an empty graph)

## Example: Scatter plots

__STEP 2__:

We want to make a scatter plot so we need to use the `geom_point` function:
```{r}
plt <- plt + geom_point()
```
Notice that all we do is "add" the required layer. Now we have something to plot:
```{r, eval=FALSE}
print(plt)
```
```{r, fig.height=2.6, fig.width=3.2, echo=FALSE}
print(plt)
```

## Example: Scatter plots

__STEP 3__:

Maybe we should improve the axis labels? To do this, we need to "add" a new labels layer using the `labs` function
```{r, eval=FALSE}
plt <- plt + labs(x = "Petal Width", y = "Petal Length")
print(plt)
```
This just adds some new information about labelling to the prexisting `ggplot` object. Now it prints with improved axis labels:
```{r, fig.height=2.6, fig.width=3.2, echo=FALSE}
plt <- plt + labs(x = "Petal Width", y = "Petal Length")
print(plt)
```

## Example: Scatter plots

__Doing it all in one go...__:

We don't have to build a plot object up in separate steps and then explicitly "print"" it to the Console. If we just want to make th eplot in one go we can do it like this:
```{r, fig.height=2.6, fig.width=3.2}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point() + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Exercise

<div id="exercise-box">
__Customising your plot__
<div id="box-text">
Repeat the example we just stepped through, but now try to customise the point colours and their size. Then see if you can make the points transparent.

Hint: the `geom_point` function is responsible for altering these features (it takes a number of different arguments). Try `?geom_point` if you can't guess the answer, looking at the **Aesthetics** section. An example of suitable output is given below.
</div>
</div>

```{r, fig.height=2.6, fig.width=3.2, echo=FALSE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Answer

```{r, fig.height=2.6, fig.width=3.2, eval=showAnswers, echo=showAnswers}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Adding more information {.build}

Q: The last graph was quite nice, but what information was missing?

A: The species identity!

```{r, fig.height=2.6, fig.width=4.7, echo=FALSE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) + 
  geom_point(size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

```{r, eval=FALSE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) + 
  geom_point(size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Exercise {.build}

<div id="exercise-box">
__Aesthetic mappings vs. arguments to `geom_`__
<div id="box-text">
You may have noticed that we can set something like "colour" in two places: the aesthetic mapping (`aes`) or via an argument to a geom (`geom_`). What happens if we set the colour in both places at once? Experiment with the iris petal length vs. petal width scatter plot example to work this out. Which one---the aesthetic mapping geom argument---has precedence?
</div>
</div>

```{r, fig.height=2.6, fig.width=3.2}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) + 
  geom_point(colour = "blue", size = 3, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Putting it all together

We want to make the following scatter plot. It shows the mean `wind` speed against mean `pressure`. The means are calculated for each combination of storm `name` and `type`. Storm type of each point is delineated by its colour.

```{r, echo = FALSE, fig.width=6}
storms %>% 
  group_by(name, type) %>%
  summarise(wind = mean(wind), pressure = mean(pressure)) %>%
  ggplot(aes(x = pressure, y = wind, col = type)) + 
    geom_point(alpha = 0.7) + 
    labs(x = "Mean pressure (mbar)", y = "Mean wind speed (mph)")
```

## Exercise {.build}

<div id="exercise-box">
__Using `dplyr` and `ggplot2` together (part 1)__
<div id="box-text">
The first step is to work out how to use `dplyr` to calcuate the mean `wind` speed and mean `pressure` for each combination of storm `name` and `type`. Do this with the pipe (`%>%`) operator, and give the resulting data the name `storms.summary`.
</div>
</div>

```{r}
storms.summary <-
  storms %>% 
  group_by(name, type) %>%
  summarise(wind = mean(wind), pressure = mean(pressure))
```

## Exercise {.build}

<div id="exercise-box">
__Using `dplyr` and `ggplot2` together (part 2)__
<div id="box-text">
The next step uses the `storms.summary` data to plot the mean `wind` speed and mean `pressure` for each name-type case. Remember to colour the points by `type`.
</div>
</div>

```{r, fig.width=6}
ggplot(storms.summary, 
       aes(x = pressure, y = wind, col = type)) + 
  geom_point(alpha = 0.7) + 
  labs(x = "Mean pressure (mbar)", y = "Mean wind speed (mph)")
```

## Exercise {.build}

<div id="exercise-box">
__Using `dplyr` and `ggplot2` together (part 3)__
<div id="box-text">
Finally, see if you can combine the solutions to part 1 and 2 into a single "piped" operation. That is, instead of storing the intermediate data in `storms.summary`, use the pipe (`%>%`) to send the data straight to `ggplot`.
</div>
</div>

```{r, fig.width=6}
storms %>% 
  group_by(name, type) %>%
  summarise(wind = mean(wind), pressure = mean(pressure)) %>%
  ggplot(aes(x = pressure, y = wind, col = type)) + 
    geom_point(alpha = 0.7) + 
    labs(x = "Mean pressure (mbar)", y = "Mean wind speed (mph)")
```


