---
title: Plotting with ggplot2
author: Dylan Childs
date: March 16, 2017
highlighter: pygments
output: 
  ioslides_presentation:
    css: extras.css
    widescreen: true
    smaller: true
---

```{r, include = FALSE}
showAnswers <- TRUE
library(dplyr)
library(ggplot2)
library(nasaweather)
library(grid)
library(knitr)

opts_chunk$set(fig.width=4.5, fig.height=3.3, warning=FALSE, message=FALSE, 
               dev.args = list(bg = 'transparent'), fig.align='center')

my_theme <- theme_gray() + theme(
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(vjust=-0.5),
  axis.title.y = element_text(vjust=+1.5)
  )
theme_set(my_theme)
```

## Before we start

Required packages and data:

- `dplyr`: obviously
- `ggplot2`: new package
- `storms.csv`: climate data for Central America (on MOLE)

**1. Load the `dplyr` and `ggplot2` packages**

```{r, message=FALSE}
library("dplyr")
library("ggplot2")
```

These should already be installed. If you see errors (you really should not by now!) you will need to install one or both of them.

__2. Read in the storms data__

Download the `storms.csv` data from MOLE. Set your working directory to the correct location and then use `read.csv` to read the data in (remember `stringsAsFactors = FALSE`), giving it the name `storms`.

Revision | How `ggplot2` works
====================================================================

## Revision: elements of `ggplot2` 

Only a few ideas needed to start using `ggplot2` effectively:

>- __layers__: We build ggplot2 objects by adding one or more layers together. We can do this in a stepwise way, only plotting the object when we are ready. Each layer needs things like data, aesthetics, geometric objects, etc
>- __aesthetics__: The word aesthetics refers to __the information__ in a plot. For example, which variables are associated with the x and y axes? We specify this using the `aes` function.
>- __geometric objects__: Geometric objects ("geoms") determine how the information is displayed. For example, will it be a scatter plot or a bar plot? We can specify geoms by adding a layer via functions beginning with `geom_`.

In practise, we build up a plot by combining different functions using the `+` operator. This has nothing to do with numeric addition!

## Revision: Scatter plots

__Doing it all in one go...__:

```{r, fig.height=2.6, fig.width=3.2}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length)) + 
  geom_point() + 
  labs(x = "Petal Width", y = "Petal Length")
```

## Making things more informative

1. Customise the plot by setting arguments of a `geom_` function.

2. Add more information by including an additional aesthetic mapping.

Option 1 takes presedence over option 2. Remember that!

```{r, fig.height=3.2, fig.width=4.8, echo=TRUE}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, colour = Species)) +
  geom_point(size = 2, alpha = 0.5) + 
  labs(x = "Petal Width", y = "Petal Length")
```


Histograms | Visualising a single variable
====================================================================

## What are histograms?

Histograms summarise the relative frequency of different values of a variable. Look at the first 56 values of `pressure` variable in `storms`:

```{r}
storms $ pressure[1:56]
```

To get a sense of how frequent different values are we can "bin" the data. Here are the frequencies of `pressure` variable values, using 8 bins:

```{r}
table(cut(storms $ pressure, breaks = 8))
```

(You don't need to rememeber this R code)

## What are histograms?

We use histograms to understand the __distribution__ of a variable. They summarise the number of observations occuring in a contiguous series of bins. Use `geom_histogram` to construct a histogram. Here is an example:

```{r}
ggplot(storms, aes(x = pressure)) + 
  geom_histogram(colour = "darkgrey", fill = "grey", binwidth=10) + 
  labs(x = "Pressure", y = "Count")  
```

## Exercise

<div id="exercise-box">
__Plotting histograms__
<div id="box-text">
Working with the `iris` dataset, construct a histogram of the ratio of petal length to petal width. See if you can make you histogram look like the one below. Hint: you can carry out the calculation with `Petal.Length` and `Petal.Width` inside `aes` (you don't have to first use `mutate` from `dplyr`)
</div>
</div>

```{r, echo=FALSE}
ggplot(iris, aes(x = Petal.Length / Petal.Width)) + 
  geom_histogram(binwidth=0.5) + 
  labs(x = "Petal Eccentricity", y = "Count")  
```

## Answer

```{r, include=showAnswers}
ggplot(iris, aes(x = Petal.Length / Petal.Width)) + 
  geom_histogram(binwidth=0.5) + 
  labs(x = "Petal Eccentricity", y = "Count")  
```

Alternative to histograms | visualising 'small' data
====================================================================

## Dot plots

We use dot plots to explore the __distribution__ of variables when we have relatively few observatins (e.g. < 100). Here is an example:

```{r}
setosa <- filter(iris, Species == "setosa")
ggplot(setosa, aes(x = Sepal.Length)) + 
  geom_dotplot(binwidth=0.1)  
```

Boxplots | Relationships between categorical and continuous data
====================================================================

## What are box and whiskers plots?

Box and whisker plots summarise the distributions of a variable at different levels of a categorical variable. Here is an example:

```{r, fig.width=3, eval=TRUE, echo=FALSE}
ggplot(iris, aes(x = Species, y = Petal.Length/Petal.Width))  + 
  geom_boxplot() + 
  labs(x = "Species", y = "Eccentricty")  
```

Each box-and-whisker shows us the group median (horizontal line) and the interquartile range ("boxes"). The vertical lines ("whiskers") are used to highlight the range of (most of) the rest of the data in each group. Points that may be outliers are plotted individually.

## Making box and whiskers plots

We can make the box and whiskers plot from the last example using `geom_boxplot` (are you starting to see the pattern)?:

```{r, fig.width=3}
ggplot(iris, aes(x = Species, y = Petal.Length/Petal.Width)) + 
  geom_boxplot() + 
  labs(x = "Species", y = "Eccentricty")  
```

## Exercise

<div id="exercise-box">
__Box and whiskers plots__
<div id="box-text">
Working with the `storms` dataset, construct a box and whiskers plot to summarise wind speed (`wind`) for each year (`year`).

Hint: you will need to convert `year` into something called a factor object using the `factor` function (we will learn more about factors next year). You can do this inside the `aes` function.
</div>
</div>

```{r, fig.width=6, echo=FALSE}
ggplot(storms, aes(x = factor(year), y = wind)) + 
  geom_boxplot() + 
  labs(x = "Year", y = "Wind Speed")  
```

## Answer

```{r, fig.width=6, include=showAnswers}
ggplot(storms, aes(x = factor(year), y = wind)) + 
  geom_boxplot() + 
  labs(x = "Year", y = "Wind Speed")  
```


Barplots | Summaries statistics for groups
====================================================================

## What are bar plots?

We typically use a barplot to summarise differences in summary statistics (like the mean) among groups. We use `geom_bar` to make barplots. When we `geom_bar` on raw data it will simply count the number of observations:

```{r, fig.height=2.5}
ggplot(storms, aes(x = factor(year))) + 
  geom_bar() + 
  labs(x = "Year", y = "Number of Storms")  
```

This is not really all that useful. Usually we use a bar plot to compare **summary statistics**.

## Using bar plots to compare means {.build}

If we want to compare a summary statistic (e.g. the mean) across groups we need to first calculate this quantity. We can do this with the `group_by` and `summarise` function from the `dplyr` package. We then use the resulting dataset to plot the means. For example:

```{r, echo=TRUE, eval=FALSE}
# step 1
pl_stats <- 
  iris %>%
  group_by(Species) %>% 
  summarise(mean_pl = mean(Petal.Length))
# step 2 
ggplot(pl_stats, aes(x = Species, y = mean_pl)) + 
  geom_bar(stat="identity") + 
  labs(y = "Mean Petal Length")
```

<div class="columns-2">
```{r, fig.width=2.5, fig.height=2, echo=FALSE}
# step 1 - use dplyr to calculate the means
pl_stats <- 
  iris %>% group_by(Species) %>% 
  summarise(mean_pl = mean(Petal.Length))
# step 2 - make the plot
ggplot(pl_stats, aes(x = Species, y = mean_pl)) + 
  geom_bar(stat="identity") + 
  labs(y = "Mean Petal Length")
```

Notice that we use `stat="identity"` as an argument to `geom_bar`. This is needed to avoid plotting the counts like last time.
</div>

## Exercise

<div id="exercise-box">
__Making a barplot of means__
<div id="box-text">
Working with the `storms` dataset, construct a bar plot to summarises the mean wind speed (`wind`) associated with storms in each year (`year`). If that was too easy, see if you can change the `fill` colour of the bars to grey.
</div>
</div>

```{r, echo=FALSE}
# step 1 - use dplyr to calculate the means
wind.means <- 
  storms %>% group_by(year) %>% 
  summarise(wind = mean(wind))
# step 2 - make the plot
ggplot(wind.means, aes(x = year, y = wind)) + 
  geom_bar(stat="identity", fill="darkgrey")
```

## Answer

```{r, include=showAnswers}
# step 1 - use dplyr to calculate the means
wind.means <- 
  storms %>% group_by(year) %>% 
  summarise(wind = mean(wind))
# step 2 - make the plot
ggplot(wind.means, aes(x = year, y = wind)) + 
  geom_bar(stat="identity", fill="darkgrey")
```

## Adding multiple layers (1)

<div class="columns-2">
```{r, fig.width=2.5, fig.height=2, echo=FALSE}
# step 1
pl_stats <- 
  iris %>%
  group_by(Species) %>% 
  summarise(mean_pl = mean(Petal.Length),
            se = sd(Petal.Length) / sqrt(n())) # <- New calculation
# step 2 
ggplot(pl_stats, 
       aes(x = Species, y = mean_pl, ymin = mean_pl - se, ymax = mean_pl + se)
       ) + 
  geom_bar(stat="identity", fill = "grey", width = 0.7) + 
  geom_errorbar(width = 0.25) + 
  labs(y = "Mean Petal Length")
```

We can build more complex figures by adding more than one layer with the `geom_` functions. For example, we should always add an error bar of some kind to summaries of means.
</div>
<br>
The __standard error__ is one option here: 

$$
\text{Standard Error} = \frac{\text{Standard Deviation}}{\sqrt{\text{Sample Size}}}
$$

We need to repeat the **dplyr**, but now include a calculation of the standard errors along with the means:

```{r, eval=FALSE}
# step 1
pl_stats <- 
  iris %>%
  group_by(Species) %>% 
  summarise(mean_pl = mean(Petal.Length),
            se = sd(Petal.Length) / sqrt(n())) # <- New calculation
```



## Using multiple geoms (2)

Once we have the two bits of information, we include these by adding two layers via two different `geom_` functions: `geom_bar` and `geom_errorbar`. We also need to define a couple new aesthetics... 

```{r, echo=TRUE, eval=FALSE}
# step 2 
ggplot(pl_stats, 
       aes(x = Species, y = mean_pl, 
           ymin = mean_pl - se, ymax = mean_pl + se)) + 
  geom_bar(stat="identity", fill = "grey", width = 0.7) + 
  geom_errorbar(width = 0.25) + 
  labs(y = "Mean Petal Length")
```


```{r, fig.width=2.5, fig.height=2, echo=FALSE}
# step 1
pl_stats <- 
  iris %>%
  group_by(Species) %>% 
  summarise(mean_pl = mean(Petal.Length),
            se = sd(Petal.Length) / sqrt(n())) # <- New calculation
# step 2 
ggplot(pl_stats, 
       aes(x = Species, y = mean_pl, ymin = mean_pl - se, ymax = mean_pl + se)
       ) + 
  geom_bar(stat="identity", fill = "grey", width = 0.7) + 
  geom_errorbar(width = 0.25) + 
  labs(y = "Mean Petal Length")
```


Congratulations! We're done!
====================================================================

## What happens next?

<br><br>

1. __There is a lecture__ tomorrow.

2. Course book...

3. DZC office hours are 12-3pm every Tuesday -- still no visitors ;(

4. Assessment after Easter


