---
title: Data frames and data wranging with dplyr
author: Dylan Childs
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  ioslides_presentation:
    css: extras.css
    smaller: no
    wide: yes
highlighter: pygments
---

```{r, echo=FALSE, warning = FALSE, message = FALSE}
library(dplyr)
library(readr)
showAnswers <- TRUE
storms <- read_csv("../data_csv/STORMS.CSV")
```

## Before we start...

**1. Go to the course Blackboard site and open today's lecture ("Block 2").**

-  Use the Google Chrome browser! Do not use Microsoft Edge.

**2. Open up RStudio from the Programs Menu on your PC.**

**3. Start a new R script using `File > New File > R Script`**

**4. Add the following line to the top of the script and run it:**

```{r, eval=FALSE}
library(readr)
```

-  Check the output in the console. If you see an error let a TA know immediately. We'll explain what this package is for later.

**5. Get your ucards out to be scanned.**

Data frames | A place for your data
======================================================================================

## Two views of data frames {.build}

The first view: __Data frames are a collection of vectors__. These may be different types (e.g. numeric, character, logical) but they must all be the same length.

<div>
Here is how to make one "by hand" using the `data.frame` function:

```{r}
mydf <- data.frame(Var1 = 1:4, Var2 = LETTERS[1:4], Var3 = c(1, 2, 4, 8))
mydf
```

You can think of each vector as corresponding to a __statistical variable__ that we want to analyse (more on that in the course book).
</div>

## Selecting vectors from data frames {.build}

We can extract a vector (e.g. called `Var1`) from a data frame (e.g. called `mydf`) in a few different ways:

<div>
- Using the vector's name and the `$` operator
```{r}
mydf $ Var1
```
- Using the vector's name and the `[[ ]]` operator
- Using the vector's position and the `[[ ]]` operator
</div>

## Selecting vectors from data frames

We can extract a vector (e.g. called `Var1`) from a data frame (e.g. called `mydf`) in a few different ways:

<div>
- Using the vector's name and the `$` operator
- Using the vector's name and the `[[ ]]` operator
```{r}
mydf[["Var1"]] # <- WE HAVE TO USE QUOTES!
```
- Using the vector's position and the `[[ ]]` operator
</div>

## Selecting vectors from data frames

We can extract a vector (e.g. called `Var1`) from a data frame (e.g. called `mydf`) in a few different ways:

<div>
- Using the vector's name and the `$` operator
- Using the vector's name and the `[[ ]]` operator
- Using the vector's position and the `[[ ]]` operator
```{r}
mydf[[1]]
```
</div>

<br><hr>

N.B. We will only use the first version in this course. Remember what `$` does. You **will** need to use it.

## Exercise {.build}

<div id="exercise-box">
__Extracting vectors from data frames__
<div id="box-text">
R has many different data sets built into it. These are stored as data frames. 

One of these is the famous `iris` data. Type this into your Console: `View(iris)`. What does the `View` function do?

How many vectors (i.e. columns) are stored in `iris`? Extract the one named `Sepal.Width` and print it to the Console.
</div>
</div>

<div>
```{r, include=showAnswers}
iris$Sepal.Width
```
</div>

## Two views of data frames {.build}

The second view: __Data frames are table-like objects__. Take another look at our simple data frame:

```{r}
mydf
```

This clearly has rows and columns. We can extract elements of the data frame using the square brackets `[ ]`:

```{r}
mydf[2, 3]
```

__[Row #, Col #]__: This extracted the element in the second row and third column. 

## Exercise {.build .smaller}

<div id="exercise-box">
__Extracting elements from data frames__
<div id="box-text">
Using the `iris` data, wee if you can use the `[,]` construct to...

1. Get the block of elements that lie in rows 2 to 3 __and__ columns 1 to 4. Hint --- you will need to use the `:` operator twice inside `[,]`

2. Get just the second row of the data set. Hint --- you **do not** need to use the `:` operator to solve this problem.
</div>
</div>

<div>
```{r, include=showAnswers}
iris[2:3, 1:4]
```
</div>

<div>
```{r, include=showAnswers}
iris[2, ]
```
</div>


__Note__: this produced a new data frame.

## Exercise {.build .smaller}

<div id="exercise-box">
__Examining a data frame__
<div id="box-text">
We've seen that `View(iris)` can be used to examine a data frame in a spreadsheet like view. There are many other functions that allow us to examine the structure of a data frame. Using the `iris` data, experiment with the `head`, `tail`, and `str` functions to see what they do.
</div>
</div>

<div>
```{r, include=showAnswers}
head(iris, n = 2)
```
</div>

<div>
```{r, include=showAnswers}
str(iris)
```
</div>

The working directory | Where R "looks" for files
======================================================================================

## What is the working directory? {.build}

A quick overview of what it is and how to work with it...

- The **working directory** is a default location where R looks for files you want to use. It is just a folder on your computer.

- If you don't set the working directory, R will do it for you. It will seldom choose the exact folder you want to use.

- YOU should set the working directory EVERY time you start R/RStudio. Setting your working directory should be the first thing you do.

- Don't do this using R code in your script. Instead, do it via the Console (hard and error prone) or using the RStudio menu system (easy and reliable).

<div>
**Set your working directory now using...** 
**`Session > Set Working Directory > Choose Directory...`**
</div>


Reading in data | A great source of confusion
====================================================================

## We need some data....

**1. Grab the "iris_morphology.xlsx" file from Blackboard**:

Look in the folder called *Data files needed for R course (semester 2)*

**2. Save this anywhere you like**

Make sure you remember where you put it!

**3. Open up your copy of "iris_morphology.xlsx" in Excel**

How many columns ("variables") and rows ("observations") are there?


## Getting data from Excel into R {.build .smaller}

Essentially a two step process

<div>
**STEP 1. Export data from Excel to a CSV file**:

- Open the "iris_morphology.xlsx" spreadsheet up in Excel

- In Excel, go to `Save As...` and choose the `Comma Separated Values (csv)` option

- Keep the default file name ("iris_morphology.csv"), but make sure the file is saved **in your working directory** (you set this earlier)
</div>
<div>
__STEP 2. Now import the CSV file into using RStudio:__

- Include the following line in your script, and send it to the Console

```{r, eval = FALSE}
iris_copy <- read_csv("iris_morphology.csv")
iris_copy
```

- What happened? Did you see any error messages? If you did, check your spelling and make sure your working directory is set to the right place.
</div>

## The other way...

**It is possible to import data in a CSV file using RStudio:**

- Via the `Import Dataset > From Text File...` menu

<BR>

**Please, please, please don't use this method!!!** 

- You have to remember do it every time you want to run your script. 

- This method is very error prone. 


Overview of **dplyr** and 'tibbles' | Getting to grips with the basics 
====================================================================

## Why use dplyr?

**dplyr** implements a __grammar of data manipulation__ to enable you manipulate data and summarise the information in a data set (e.g. group means).

__Advantages of using **dplyr**__
 
- Provides a consistent framework for data manipulation
- Designed to work well with the `ggplot2` plotting system
- Fast compared to many base R functions 
- Allows you to work with data stored in many ways (e.g. in a database)

## Five key verbs {.build} 

<div>
__**dplyr** has five main "verbs" (i.e. functions):__

- `select`: Extract a subset of variables
- `filter`: Extract a subset of rows
- `arrange`: Reorder rows 
- `mutate`: Construct new variables 
- `summarise`: Calculate information about groups

We'll also explore a few more useful functions such as `slice`, `rename`, `transmute`, and `group_by`. There are many others...
</div>

<div>
<br>

## Five key verbs
__It is helpful to classify the verbs according to what they work on:__

- observations (rows): `filter` & `slice` & `arrange`
- variables (columns): `select` & `rename` & `mutate`
- summarise: `summarise` (or `summarize`)

(This classification only works if your data are __tidy__, i.e. there is one observation per row and one column per variable. Make sure you read about this idea in the course book)
</div>

## Before we start {.build .smaller}

<div>
**Required packages, data and scripts:**

- The **readr** and **dplyr** packages
- The data file: `CENTRAL_AMERICA_STORMS.CSV` (on Blackboard)
- A template R script: `block-2-template.R` (on Blackboard)
</div>

<div>
**Make sure you carry out these steps:**

- Set your working directory.

- Run the code to load the __dplyr__ and __readr__ packages:

```{r, eval=FALSE}
library(readr)
library(dplyr)
```

- Run the code to read the Central American storms data into R:

```{r, eval=FALSE}
storms <- read_csv("CENTRAL_AMERICA_STORMS.CSV")
```

**Check for errors at every step!**
</div>

## What is storms? {.smaller}

Use `View` to examine the `storms` object:

```{r, echo=FALSE}
print(storms, n=12)
```

...`storms` is something called a 'tibble' (technically a `tbl` object). This is the tidyverse version of a data frame.

## Making and using tibbles {.smaller}

A `tbl` object ("tibble") is essentially a special kind of data frame that belongs to the tidyverse. It mostly works the same as a data frame, but with a few small differences... 

e.g. compact printing---here is what happens when we print a **standard data frame**:

```{r}
# print the data frame version
iris
```

## Making and using tibbles {.smaller}

A `tbl` object ("tibble") is essentially a special kind of data frame that belongs to the tidyverse. It mostly works the same as a data frame, but with a few small differences... 

e.g. compact printing---here is what happens when we print the equivalent **tibble**:

```{r}
# convert to tibble and print this version
as_tibble(iris)
```

## (Convert ``iris` to tibbles (optional)

Next in your script, add these lines to convert `iris` to a tibble...

```{r}
iris <- as_tibble(iris)
```

You don't have to do this (**dplyr** is fine with normal data frames) but it will ensure your output matches the presentation. 

The __readr__ package always makes a tibble so there's no need to manually convert `storms` to a tibble.

## Looking at your data {.smaller}

In addition to printing a `tbl` or `data.frame`, we can use the `glimpse` function to obtain different summary information about variables:

```{r}
glimpse(storms)
```

This is similar to `str` --- the `glimpse` function tells us what variables are in `storm` as well as the type of each variable.

Using `select` | Extracting a subset of variables 
====================================================================

## Basic usage {.smaller}

We use `select` to extract a subset of variables for further analysis. Using `select` looks like this:

```{r, eval=FALSE}
select(<data>, <variable-1>, <variable-2>, ...)
```

__Arguments__

- `<data>`: a `data.frame` or `tbl` object 
- `<variable-X>`: names of a variable in `<data>`

<div>
__Comments__

- You select as many variable as you want by including more than two variables after the data. That is what the `...` denotes.
</div>


## Exercise {.build}

<div id="exercise-box">
__Selecting two variables__
<div id="box-text">
Use the `select` function with the `storms` data set to make a new data set containing only `name` and `year`. Assign this new data set a name, and then check that it contains the right variables using the `glimpse` function.
</div>
</div>

```{r, include=showAnswers}
storms_simple <- select(storms, name, year)
glimpse(storms_simple)
```

## Selecting & dropping variable ranges {.build}

The `select` function makes selecting/removing groups of variables easy:

- Use `:` to select a sequence of variables 
- Use `-` to drop a sequence of variables 

The sequence can be specified using numbers (for position) or names.

<div>
__Usage:__

```{r, eval=FALSE}
# a range of variable to keep
select(<data>, <variable-1>:<variable-2>)
# drop a variable
select(<data>, -<variable-1>)
```
</div>

<div>
## Example:

```{r}
iris_fewer <- select(iris, Petal.Length:Species)
iris_fewer
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Selecting a range of variables__
<div id="box-text">
Use the `select` function with the `storms` data set to select __just__ the variables `name`, `year` and `month` variables.
</div>
</div>

```{r, include=showAnswers}
storms_fewer <- select(storms, name:month)
glimpse(storms_fewer)
```

```{r, include=showAnswers}
# alternatively
x <- select(storms, -(day:seasday))
```

## Helper functions {.build .smaller}

There are several helper functions that work with `select` to simplify common variable selection tasks:

- `starts_with("xyz")`: every name that starts with `"xyz"`
- `ends_with("xyz")`: every name that ends with `"xyz"`
- `contains("xyz")`: every name that contains `"xyz"`
- `matches("xyz")`: every name that matches `"xyz"`
- `one_of(names)`: every name that appears in `names` (character vector).

<div>
__Usage:__

```{r, eval=FALSE}
select(<data>, <help_func>("xyz"))
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_petal <- select(iris, starts_with("Petal"))
glimpse(iris_petal)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Using select with helper functions__
<div id="box-text">
Use the `select` function with the `storms` data set to create a new data set containing just the `lat` and `long` variables. Do this using the `starts_with` helper function inside select. 
</div>
</div>

```{r, include=showAnswers}
storms_fewer <- select(storms, starts_with("l"))
glimpse(storms_fewer)
```

Using `select` and `rename` | Renaming variables
====================================================================

## Renaming while selecting {.smaller}

We can use `select` to rename variables as we select them using a `<new-name> = <old-name>` construct.

<div>
__Usage:__

```{r, eval=FALSE}
select(<data>, 
       <new-name-1> = <old-name-1>, 
       <new-name-2> = <old-name-2>, ...)
```
</div>

<div>
__Example:__

```{r, eval=TRUE}
iris_select_rename <- select(iris, 
                             petal_length = Petal.Length,
                             petal_width  = Petal.Width)
glimpse(iris_select_rename)
```
</div>

## Renaming without selecting {.smaller}

Use `rename` to rename variables while keeping all variables using the `<new-name> = <old-name>` construct.

<div>
__Usage:__

```{r, eval=FALSE}
rename(<data>, 
       <new-name-1> = <old-name-1>, 
       <new-name-2> = <old-name-2>, ...)
```
</div>

<div>
__Example:__

```{r}
iris_renamed <- rename(iris, 
                       petal_length = Petal.Length,
                       petal_width  = Petal.Width)
glimpse(iris_renamed)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Renaming__
<div id="box-text">
Extract just the `lat` and `long` columns from the storms data set and rename these as `latitude` and `longitude` respectively.

</div>
</div>

```{r, include=showAnswers}
storms_renamed <- select(storms, 
                         latitude = lat, 
                         longitude = long)
glimpse(storms_renamed)
```

## Part 2 --- Before we start {.smaller}

<div>
**Required packages, data and scripts:**

- The **readr** and **dplyr** packages
- The data file: `CENTRAL_AMERICA_STORMS.CSV` (on Blackboard)
- A template R script: `block-2-template.R` (on Blackboard)
</div>

<div>
**Make sure you carry out these steps:**

- Set your working directory.

- Run the code to load the __dplyr__ and __readr__ packages:

```{r, eval=FALSE}
library(readr)
library(dplyr)
```

- Run the code to read the Central American storms data into R:

```{r, eval=FALSE}
storms <- read_csv("CENTRAL_AMERICA_STORMS.CSV")
```

**Check for errors at every step!**
</div>


Using `mutate` | Making new variables 
====================================================================

## Basic usage {.smaller}

<div>
We use `mutate` to __add variables__ for further analysis. This is how we use `mutate`:

```{r, eval=FALSE}
mutate(<data>, <new-var-name> = <expression>, ...)
```
</div>

<div>
__Arguments__

- `<data>`: a `data.frame` or `tbl` object 
- `<new-var-name>`: name of a new variable to create
- `<expression>`: an R expression that references variables in `<data>`
</div>

<div>
__Comments__

- The `<expression>` which appears on the right hand side of the `=` can be any valid R expression that uses variables in `<data>`. 

- You may use more than one `<new-var-name> = <expression>` at a time if you need to construct several new variables.
</div>

## Exercise {.build}

<div id="exercise-box">
__Making a new variable__
<div id="box-text">
Use the `mutate` function with the `iris` data set to make a new variable which is the petal area, $(Area = Length \times Width)$.
</div>
</div>

```{r, include=showAnswers}
iris_area <- mutate(iris, 
                    Petal.Area = Petal.Length * Petal.Width)
glimpse(iris_area)
```

## Multiple calculations {.build}

<div>
We can add more than one variable at a time using `mutate`. Each new variable can also use one or more variables created in a previous step.
</div>

<div>
__Usage:__

```{r, eval=FALSE}
mutate(<data>, 
       <new-var-name-1> = <expression-1>, 
       <new-var-name-2> = <expression-2-using-new-var-1>)
```
</div>

<div>
__Example:__

```{r, eval=FALSE}
iris_new_vars <- 
  mutate(iris, 
         Sepal.Eccentricity = Sepal.Length / Sepal.Width,
         Petal.Eccentricity = Petal.Length / Petal.Width,
         Eccentricity.Diff  = Sepal.Eccentricity - Petal.Eccentricity)
```
</div>

## Exercise {.build}

<div id="exercise-box">
__Making several new variables__
<div id="box-text">
Use the `mutate` function with the `iris` data set to make two new area variables, one for petal and one for sepal. Create a third variable which is the ratio of the petal and sepal areas. 

Do all of this in one call to `mutate`, i.e. use `mutate` only once to do all of this.
</div>
</div>

```{r, include=showAnswers}
iris_ratio <- mutate(iris, 
                     Sepal.Area = Sepal.Length * Sepal.Width,
                     Petal.Area = Petal.Length * Petal.Width,
                     PS.Area.Ratio = Petal.Area / Sepal.Area)
glimpse(iris_ratio)
```

Using `filter` | Selecting subsets of observations 
====================================================================

## Basic usage {.smaller}

<div>
We use `filter` to select a subset of rows for further analysis, based on the result(s) of one or more logical comparisons. Using `filter` looks like this:

```{r, eval=FALSE}
filter(<data>, <expression>)
```
</div>

<div>
__Arguments__

- `<data>`: a `data.frame` or `tbl` object 
- `<expression>`: an R expression that implements a logical comparison using variables in `<data>`
</div>

<div>
__Comments__

- The `<expression>` can be any valid R expression that uses variables in `<data>` and returns a logical vector of `TRUE` / `FALSE` values. 

- The `<expression>` typically uses a combination of __relational__ (e.g. `<` and `==`) and __logical__ (e.g. `&` and `|`) operators.
</div>

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on one variable__
<div id="box-text">
Use the `filter` function with the `storms` data set to create a new data set containing just the observations associated with storms classified as Hurricanes.

Hint: use `glimpse` to remind yourself of the variable names in `storms`. You need to work out which one contains information about the storm category.
</div>
</div>

```{r, include=showAnswers}
filter(storms, type == "Hurricane")
```

## Exercise {.build}

<div id="exercise-box">
__Subsetting observations on more than one variable__
<div id="box-text">
Repeat the last exercise, but now extract the observations associated with Hurricanes that took place in 1997 or later.
</div>
</div>

```{r, include=showAnswers}
filter(storms, 
       type == "Hurricane", year >= 1997) 

# or use: filter(storms, type == "Hurricane", year >= 1997)
```


Using `arrange` | Reordering observations 
====================================================================

## Basic usage {.smaller}

<div>
We use `arrange` to reorder the rows of our data set. This can help us see associations among our variables. Using `arrange` looks like this:

```{r, eval=FALSE}
arrange(<data>, <variable-1>, <variable-2>, ...)
```
</div>

<div>
__Arguments__

- `<data>`: a `data.frame` or `tbl` object 
- `<variable-X>`: names of variables in `<data>`
</div>

<div>
__Comments__

- The order of sorting corresponds to the order that variables appear in `arrange`, meaning `<data>` is sorted according to `<variable-1>`, then `<variable-2>`, then `<variable-3>`, etc. 

- The sort order is from smallest to largest (ascending). If you want to reverse the sort order to go from largest to smallest (descending) use `desc(<variable-X>)`.
</div>

## Exercise {.build}

<div id="exercise-box">
__Reording observations__
<div id="box-text">
Use the `arrange` function to reorder the observations in the `storms` data set, according to the pressure variable. Store the resulting data set and use then use the `View` function to examine it. What can you say about the association between atmospheric pressure and storm category?
</div>
</div>

```{r, eval=FALSE}
storm_sort <- arrange(storms, pressure)
View(storm_sort)
```

Using `summarise` | Calculating summaries of variables 
====================================================================

## Basic usage {.smaller}

<div>
We use summarise to __calculate summary statistics__ for further analysis. This is how to use `summarise`:

```{r, eval=FALSE}
summarise(<data>, <summary-var-name> = <expression>, ...)
```
</div>

<div>
__Arguments__

- `<data>`: a `data.frame` or `tbl` object 
- `<summary-var-name>`: name of your summary variable
- `<expression>`: an R expression that references variables in `<data>` and returns to __a single value__
</div>

<div>
__Comments__

- The `<expression>` which appears on the right hand side of the `=` can be any valid R expression that uses variables in `<data>`. However, `<expression>` should return __a single value__.

- Although `summarise` looks a little like `mutate`, it is designed to construct a completely new data set containing summaries of one or more variables.

- You may use more than one `<summary-var-name> = <expression>` at a time if you need to construct several summaries.
</div>

## Exercise {.build}

<div id="exercise-box">
__Calculating the mean of two variables__
<div id="box-text">
Use the `summarise` function with the `iris` data set to calculate the mean sepal length and the mean sepal width.

Hint: You need to work out which R function calculates a mean. The clue is in the name.
</div>
</div>

```{r, include=showAnswers}
summarise(iris, 
          mean_sl = mean(Sepal.Length), 
          mean_sw = mean(Sepal.Width))
```

## Exercise {.build}

<div id="exercise-box">
__Calculating a more complex summary of a variable__
<div id="box-text">
Use the `summarise` function with the `iris` data set to calculate the mean __area__ of sepals.
</div>
</div>

```{r, include=showAnswers}
summarise(iris, 
          mean_sl = mean(Sepal.Length * Sepal.Width))
```

```{r, eval=FALSE, include=showAnswers}
# THIS IS WRONG!
summarise(iris, 
          mean_sl = mean(Sepal.Length) * mean(Sepal.Width))
```

Using `group_by` | Making summaries for groups of observations
====================================================================

## Basic usage {.smaller}

<div>
We use `group_by` to add grouping information to a data frame or tibble. This is how we use `group_by`:

```{r, eval=FALSE}
group_by(<data>, <grouping-var-1>, <grouping-var-2>, ...)
```
</div>

<div>
__Arguments__

- `<data>`: a `data.frame` or `tbl` object 
- `<grouping-var-X>`: name of grouping variable(s) in `<data>`
</div>

<div>
__Comments__

- The `group_by` does not do anything other than add grouping information to a `tbl`. It is only useful when used with `summarise` or `mutate`.

- Using `group_by` with `summarise` enables us to calculate numerical summaries on a _per group_ basis.

- Using `group_by` with `mutate` enables us to add variables on a _per group_ basis.
</div>

## Exercise {.build .smaller}

<div id="exercise-box">
__Using `group_by` to calculate group-specific means__
<div id="box-text">
Use the `group_by` function and the `summarise` functions with the `storms` data set to calculate the mean wind speed associated with each storm __type__.

Hint: This is a two step exercise: 1) Use `group_by` to add some information to storms, remembering to assign the result a name; 2) These use `summarise` on this new data set.
</div>
</div>

```{r, include=showAnswers}
# 1. make a grouped tibble
storms_grouped <- group_by(storms, type)
# 2. use summarise on the grouped data
summarise(storms_grouped, mean_wind = mean(wind))
```

## Exercise {.build .smaller}

<div id="exercise-box">
__Using `group_by` to group by more than one variable__
<div id="box-text">
Use the `group_by` and `summarise` functions with the `storms` data set to calculate the mean and maximum wind speed associated with **each combination of month and year**. Use `View` to examine the result. Which month in which year saw the largest maximum wind speed? 
</div>
</div>

<div>
```{r, include=showAnswers, eval=FALSE}
# 1. make a grouped tibble
storms_grouped <- group_by(storms, year, month)
# 2. use summarise on the grouped data
summarise(storms_grouped, 
          mean_speed = mean(wind), max_speed = max(wind))
```
Next slide shows what we want to make.
</div>

## Exercise {.smaller}

Aiming for this...

```{r, echo=FALSE}
group_by(storms, year, month) %>% 
  summarise(mean_speed = mean(wind), max_speed = max(wind)) %>% 
  print(n = 18)
```

## Exercise {.smaller .build}

<div id="exercise-box">
__Using `group_by` and `mutate`__
<div id="box-text">
The `group_by` function works with any **dplyr** verb that operates on variables (columns). Use the `group_by` and `mutate` functions with the `iris` data set to calculate a "mean centred" version of sepal length. A centred variable is just one that has had its overall mean subtracted from every value. 

Do you understand the different behaviour of `summarise` and `mutate` when used alongside `group_by`?
</div>
</div>

```{r, include=showAnswers}
# 1. group iris by species identity
iris_grouped <- group_by(iris, Species)
# 2. use mutate on the grouped data
mutate(iris_grouped, sl_centred = Sepal.Length - mean(Sepal.Length))
```

We're finished with block 2 | Next... piping and making figures
====================================================================

