---
layout: default
title: Exercises - functions and vectors 
---

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=7, fig.height=8, warning=TRUE, error=TRUE, prompt=FALSE)
```

### Question XXXX

Take the natural log of 5 and round this to 3 decimal places. Do this with just one line of R code, i.e. make sure you do the whole calculation in one step.

#### Answer

The `log` function takes the natural log of a number (or more accurately, it takes the natural log of the elements of a numeric vector). This function only needs a single argument (called `x`), which is the numeric vector. The `round` function rounds the elements of a numeric vector to a user-specified number of decimal places. This takes two arguments: `x` is the numeric vector, and `digits` is an integer specifying the decimal places. In order to use these two functions together you have to remember to use function nesting, like this:

```{r}
round(x = log(5), digits = 3)
```

You could also have done this without naming either of the arguments...

```{r}
round(log(5), 3)
```

...though it is generally safer to name all but the first argument...

```{r}
round(log(5), digits = 3)
```
